---
title: "Analysis scenith data"
author: "Léonard Dubois, Rodrigo Formiga and Harry Sokol"
date: "2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(phyloseq)
library(microViz)
library(viridis)
library(vegan)
library(ggsignif)
library(rstatix)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)

rm(list = ls())
gc()

colormap_group <- c("#b2df8a", "#a6cee3", "#1f78b4") 
names(colormap_group) <- c("HD", "before_FMT", "after_FMT") 

```


# Import data

```{r}

mpa4_table <- read_tsv(file = "../data/profile_mpa4_vJun23_FMT_Cdiff.tsv", col_names = T, show_col_types = F)
scenith_dt <- read_tsv(file = "../data/scenith_data.tsv", col_names = T, show_col_types = F)
md_mpa4_group <- tibble(sample = colnames(mpa4_table)[-1])
md_mpa4_group <- md_mpa4_group %>% 
    left_join(scenith_dt %>% select(sample, group)) %>% 
    mutate(group = ifelse(is.na(group), "HD", group)) %>% distinct()

rname <- mpa4_table$clade
mpa4_table <- as.matrix(mpa4_table %>% select(-clade))
rownames(mpa4_table) <- rname

mpa4_table_sgb <- mpa4_table[ grepl("t__SGB", rownames(mpa4_table) ), ]
mpa4_table_spc <- mpa4_table[ grepl("s__[A-Za-z_]+$", rownames(mpa4_table) ), ]


pairwise_FMT_dt <- read_tsv(file = "../data/FMT_pairing.tsv", col_names = T, show_col_types = F)
pairwise_FMT_dt_shotgun <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics") %>% 
    mutate(FMT_group = paste0("FMT_grp_", row_number())) %>% 
    pivot_longer(cols = c("donor", "before_FMT", "after_FMT"), values_to = "sample")
md_mpa4_group <- md_mpa4_group %>% left_join(pairwise_FMT_dt_shotgun %>% select(sample, FMT_group))

```

Change here to run the chunks below considering Speceis or SGB level

```{r}
mpa4_table_used <- mpa4_table_spc

```



# Alpha div

```{r}
spec_richness_0_05 <- tibble(sample =  colnames(mpa4_table_used),
                             richness = apply(mpa4_table_used, 2, function(x) sum(x > 0.05))) %>% 
    left_join(md_mpa4_group) %>%
    mutate(group = factor(group, levels = c("HD", "before_FMT", "after_FMT")))


ggplot(data = spec_richness_0_05, 
       aes(x = group, y = richness, color = group)) + 
    geom_boxplot(outlier.shape = NA) + 
    geom_jitter(height = 0, size = 2) + 
    theme_minimal() +
    ggpubr::stat_pwc( tip.length = 0, p.adjust.method = "BH") + 
    scale_color_manual(values = colormap_group) +
    xlab("") + ylab("Species richness (relative abundance>=0.05%)") +
    theme(panel.grid.major.x = element_blank(), 
          legend.position = "none" )

```


# Beta diversity


prepare dist data

```{r}

dt <- mpa4_table_used[ , apply(mpa4_table_used, 2, function(x) max(x) < 95) ]
dt <- dt[rowSums(dt) != 0, ]

# or better  use direct asv relab
dist_beta <- vegdist(t(dt), method = "aitchison", pseudocount = 1e-10)
# Jaccard is also interesting
#dist_beta <- vegdist(t(dt), method = "jaccard", pseudocount = 1e-10)

md <- md_mpa4_group %>% select(sample, group) %>% distinct()
pco_dt <- ecodist::pco(dist_beta)
plot_dt <- as.data.frame(pco_dt$vectors[, 1:2])
plot_dt$sample <- rownames(as.matrix(dist_beta))
plot_dt <- plot_dt %>%  left_join(md)

set.seed(12345)
adonis_res <- adonis2(dist_beta ~ group, data = md, permutations = 1e3)
res_permutest <- permutest(betadisper(d = dist_beta, group = md$group), permutations = 1e3)

p <- plot_dt %>%
    ggplot(data=., aes(x=X1, y=X2) ) + 
    geom_point(size = 2, alpha = 0.9, pch = 21,
               aes(fill = group)) +
    stat_ellipse(aes(color = group)) +
    theme_minimal() + 
    scale_fill_manual(values = colormap_group) + 
    scale_color_manual(values = colormap_group) +
    xlab(paste("PC1", round(100*(pco_dt$values[1] / sum(pco_dt$values)), digits = 2), "%")) + 
    ylab(paste("PC2", round(100*(pco_dt$values[2] / sum(pco_dt$values)), digits = 2), "%")) + 
    #coord_equal() +
    theme(legend.position = "right", 
          legend.direction = "vertical") +
    labs(fill = "", color = "") + 
    # adjust title
    ggtitle("Beta diversity - Aitchison distance", 
            subtitle = paste0("adonis2 R2 = ", as.character(round(adonis_res$R2[1] * 100, 2)),
                              "%;P=", as.character(round(adonis_res$`Pr(>F)`[1], 4)), 
                              "\npermutest P=", as.character(round(res_permutest$tab["Groups", "Pr(>F)"], 4)) ))

```


# Heatmap of composition

Define top species by group
```{r}
TOP_N = 15

rownames(mpa4_table_used) <- rownames(mpa4_table_used) %>% str_remove_all(pattern = ".*s__") %>%
    str_replace_all(pattern = "\\|t__", replacement = " ")

top_spc_grp <- list()
# order is important here to highlights overlap is any
for(grp in c("before_FMT", "HD", "after_FMT")){
    #grp  = "before_FMT"
    tmp <- mpa4_table_used[, md_mpa4_group%>% filter(group == grp) %>% pull(sample)]
    percentile90 <- apply(tmp, 1, function(x) quantile(x, 0.9))
    if(!is.null( unlist(top_spc_grp))){
        percentile90 <- percentile90[!names(percentile90) %in% unlist(top_spc_grp)]
    }
    top_spc <- names(sort(percentile90, decreasing = T))[1:TOP_N]
    top_spc_grp[[grp]] <- top_spc
}

row_split_dt <- rep(paste0("top species in ", names(top_spc_grp)), each = TOP_N) 
names(row_split_dt) <- unlist(top_spc_grp)

```



```{r}

relab_dt <- mpa4_table_used[names(row_split_dt), ]

# relab_dt <- relab_dt[apply(mpa4_table_used, 1, function(x) sum(x > 0.05)/ncol(mpa4_table_used)) > 0.4, ]

col_fun_HM <- colorRamp2(c("white", brewer.pal(n = 7, name = "RdYlBu"), "black"),
                         breaks = c(0, 0.01, 0.1, 1, 5, 10, 20, 50, 100))

col_split_dt <- md_mpa4_group %>% select(sample, group) %>% distinct() %>% pull(group)
names(col_split_dt) <- md_mpa4_group %>% select(sample, group) %>% distinct() %>% pull(sample)

h <- Heatmap(matrix = relab_dt, 
             col = col_fun_HM,
             
             width = unit(0.40, "cm") * ncol(relab_dt), 
             height = unit(0.40, "cm") * nrow(relab_dt),
             
             row_split = factor(row_split_dt, levels = c("top species in before_FMT", "top species in HD", "top species in after_FMT")),
             
             cluster_rows = T,
             clustering_distance_rows = "spearman", clustering_method_rows = "ward.D2",
             clustering_distance_columns  = "spearman", clustering_method_columns = "ward.D2",

             cluster_columns = T, 
             show_column_names = T, 
             column_split = factor(col_split_dt, levels = c("before_FMT", "HD", "after_FMT")), 
             show_row_dend = F, show_column_dend = F,
             
              cluster_row_slices = FALSE,
             cluster_column_slices = FALSE,
             
             row_title_side = "right",
             row_title_gp = gpar(fontsize = 10),
             column_title_gp = gpar(fontsize = 10),
             border = T,                
             column_gap = unit(3, "mm"),
             row_names_gp = gpar(fontsize = 8, fontface= "italic"),
             column_names_gp = gpar(fontsize = 8),
             
             rect_gp = gpar(col = "grey", lwd = 0.5),
             show_heatmap_legend =T,
             column_title_side = "bottom", 
             column_title_rot = 0,
             row_names_side = "left", row_dend_side = "right",
             heatmap_legend_param = list(title = "Relative abundance", 
                                         direction = "vertical", 
                                         title_gp = gpar(fontsize = 10), 
                                         legend_gp = gpar(fontsize = 10),
                                         legend_width = unit(6, "cm"),
                                         break_dist = 1,
                                         at = c(0, 0.01, 0.1, 1, 10, 50, 100)))
h


```



# OLD Correlation


## Correlation microbiome-SCENITH

### First, taking everything as one for more power, not split before after FMT

```{r}
sample_to_keep <- intersect(colnames(mpa4_table_used), scenith_dt$sample)

res <- lapply(unique(scenith_dt$type_measure), function(x){
    # loop on measure type
    #x = "Mitochondrial dependence"
    tmp_scenith <- scenith_dt %>% filter(type_measure == x, sample %in% sample_to_keep)
    tmp_mpa <- mpa4_table_used[, sample_to_keep]
    tmp_mpa <- mpa4_table_used[, tmp_scenith$sample] # change order
    tmp_mpa <- tmp_mpa[apply(tmp_mpa, 1, function(x) sum(x > 0.05) / ncol(tmp_mpa)) > 0.4, ]
    # loop on cell type/ scenith column
    res_corr <- lapply(colnames(tmp_scenith)[2:13], function(cell_type){
        #cell_type = "B cells" 
        # loop on SGB
        lapply(rownames(tmp_mpa), function(sgb){
            res <- cor.test(tmp_mpa[sgb, ], unlist(tmp_scenith[ , cell_type]), method = "spearman" )
            return(tibble(sgb = sgb, 
                          cell_type = cell_type, 
                          corr = res$estimate, 
                          p = res$p.value, 
                          measure = x))
        }) %>% do.call(., what = rbind)
    }) %>% do.call(., what = rbind)
    res_corr$padj <- p.adjust(res_corr$p, method = "BH")
    return(res_corr)
}) %>% do.call(., what = rbind)

res$sgb <- res$sgb %>% str_remove_all(pattern = ".*s__") %>% str_replace_all(pattern = "\\|t__", replacement = " ")

res %>% filter(padj < 0.1)

```


```{r}
MEASURE <- "Mitochondrial dependence"
#MEASURE <- "Glucose dependence"
#MEASURE <- "PS/ATP"


corr_dt <- res %>% filter(measure == MEASURE) %>% 
    select(sgb, cell_type, corr) %>% 
    pivot_wider(names_from = cell_type, values_from = corr) %>%
    as.data.frame()
rownames(corr_dt) <- corr_dt$sgb
corr_dt <- corr_dt %>% select(-sgb)


pval_dt <- res %>% filter(measure == MEASURE) %>% 
    select(sgb, cell_type, padj) %>% 
    pivot_wider(names_from = cell_type, values_from = padj) %>%
    as.data.frame()
rownames(pval_dt) <- pval_dt$sgb
pval_dt <- pval_dt %>% select(-sgb)

col_fun_HM <- colorRamp2(c("blue", "white", "red"),
                             breaks = c(1, 0, -1))
h <- Heatmap(matrix = corr_dt, 
            col = col_fun_HM,
            na_col = "white",
                 
            width = unit(0.30, "cm") * ncol(corr_dt), 
            height = unit(0.30, "cm") * nrow(corr_dt),
            
            column_title = MEASURE,
            column_title_gp = gpar(fontsize = 12),
            column_title_side = "top", 
            cluster_rows = T, 
            show_row_dend = F,
            cluster_columns = T, 
            show_column_names = T,
            show_row_names = T, row_title = NULL,
            
            border = T,                
            column_gap = unit(1, "mm"), 
            row_names_gp = gpar(fontsize = 9, fontface= "italic"),     
            column_names_gp = gpar(fontsize = 9),
            
            rect_gp = gpar(col = "white", lwd = 0.5),
            show_heatmap_legend =T,
            row_names_side = "left", 
            
            cell_fun = function(j, i, x, y, width, height, fill) {
                if(!is.na(pval_dt[i, j] )){
                    if(pval_dt[i, j] < 0.1){
                        color_star <- ifelse(abs(corr_dt[i, j]) > 0.5, "white", "black")
                        grid.text("*", x, y -height/3, gp = gpar(fontsize = 12, col = color_star))
                    }
                }
            })

draw(h, heatmap_legend_side = "right", )

```






### Then split by group before and after FMT

Splitting by group before and after FMT
actually makes no sense to look for correlation before FMT as the alpha diversity is very low
And the species present are actually not so prevalent afterwards 

```{r}

sample_to_keep <- intersect(colnames(mpa4_table_used), scenith_dt$sample)
sample_target <- md_mpa4_group %>% filter(group == "after_FMT", sample %in% sample_to_keep) %>% pull(sample)


res_after_FMT <- lapply(unique(scenith_dt$type_measure), function(measure){
    # loop on measure type
    # measure = "Mitochondrial dependence"
    tmp_scenith <- scenith_dt %>% filter(type_measure == measure, sample %in% sample_target)
    tmp_mpa <- mpa4_table_used[ , sample_target]
    tmp_mpa <- mpa4_table_used[ , tmp_scenith$sample] # change order
    tmp_mpa <- tmp_mpa[apply(tmp_mpa, 1, function(x) sum(x > 0.05) / ncol(tmp_mpa)) > 0.9, ]
    # loop on cell type/ scenith column
    res_corr <- lapply(colnames(tmp_scenith)[2:13], function(cell_type){
        #cell_type = "B cells" 
        # loop on SGB
        lapply(rownames(tmp_mpa), function(sgb){
            res <- cor.test(tmp_mpa[sgb, ], 
                            tmp_scenith %>% filter(sample %in% sample_target) %>% pull(cell_type), 
                            method = "spearman",  )
            return(tibble(sgb = sgb, 
                          cell_type = cell_type, 
                          corr = res$estimate, 
                          p = res$p.value, 
                          measure = measure))
        }) %>% do.call(., what = rbind)
    }) %>% do.call(., what = rbind)
    
    res_corr$padj <- p.adjust(res_corr$p, method = "BH")
    return(res_corr)
}) %>% do.call(., what = rbind)

res_after_FMT %>% filter(padj < 0.1)
# res_after_FMT %>% filter(p < 0.05)

# only 1 correlation, visualization would be useless


```



