---
title: "Analysis scenith data"
author: "Léonard Dubois, Rodrigo Formiga and Harry Sokol"
date: "2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(phyloseq)
library(microViz)
library(viridis)
library(vegan)
library(ggsignif)
library(rstatix)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)

rm(list = ls())
gc()


colormap_group <- c("#b2df8a", "#a6cee3", "#1f78b4") 
names(colormap_group) <- c("HD", "before_FMT", "after_FMT") 

colormap_enrich <- c('#b35806','#f1a340','#fee0b6', "lightgrey", '#d8daeb', '#998ec3','#542788')
names(colormap_enrich) <- c("restored over", "restored normal", "restored partial", "NS", "depleted partial", "depleted normal", "depleted over")
     

comparisons_grp_rm <- c("pre_post_FMT" = "donor", "pre_donor" = "after_FMT", "post_donor" = "before_FMT")
pairwise_FMT_dt <- read_tsv(file = "../data/FMT_pairing.tsv", col_names = T, show_col_types = F)  %>% select(-comment) %>%
    mutate(FMT_group = paste0("FMT_grp_", row_number())) %>% 
    pivot_longer(cols = c("donor", "before_FMT", "after_FMT"), values_to = "sample")

```



# Plan

For each type of data (Microbiome, metabo, scenigth, Humann) look for differences and evolution between:
- Pre FMT and Post FMT
- Pre FMT and Healthy donors
- Healthy donors and post FMT
And intersect between these test

Measure the shift (pre post FMT) compared to the mean of Healthy donors


Then look for engraftment of species (% of abundance of species coming from donor)

Build matrix of significant measurements (bacteria, metabolites, SCENITH ) to look for correlation

```{r}

all_measure_mat <- NULL
all_measure_type <- tibble()

```


# Microbiota data


## DATA

```{r}
mpa4_table <- read_tsv(file = "../data/profile_mpa4_vJun23_FMT_Cdiff.tsv", col_names = T, show_col_types = F)

rname <- mpa4_table$clade
mpa4_table <- as.matrix(mpa4_table %>% select(-clade))
rownames(mpa4_table) <- rname

#mpa4_table_spc <- mpa4_table[ grepl("s__[A-Za-z_]+$", rownames(mpa4_table) ), ]
mpa4_table_spc <- mpa4_table[ grepl("t__SGB", rownames(mpa4_table) ), ]
rownames(mpa4_table_spc) <- str_remove(string = rownames(mpa4_table_spc), ".*s__")
```


## tests

```{r}

spc_keep <- apply(mpa4_table_spc, 1, function(x) sum(x > 0.05) / ncol(mpa4_table_spc) >= 0.3) # prevalence threshold
mpa_dt_light <- mpa4_table_spc[spc_keep , ]


mean_relab <- lapply(rownames(mpa_dt_light), function(x){
    tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics") %>%
        distinct(name, sample) %>%
        filter( sample %in% colnames(mpa_dt_light)) 
    tmp$spc_abundance <- mpa_dt_light[x, tmp$sample]
    res <- tmp %>% group_by(name) %>% 
        summarise(med_relab = mean(spc_abundance)) %>%
        mutate(spc = x)
    return(res)
}) %>% do.call(., what= rbind)


res_microbiota <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")  %>% 
                                 filter(name != comparisons_grp_rm[comparison], sample %in% colnames(mpa4_table_spc)) 
                             tmp <- tmp %>% group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                             mpa_dt <- mpa_dt_light[, tmp$sample]
                           
                             res_test <- lapply(rownames(mpa_dt), function(x){
                                 tmp$spc_abundance <- mpa_dt[x, ]
                                 if(var(tmp$spc_abundance) == 0){ return(NULL)}
                                 test <- wilcox_test(data = tmp, formula = spc_abundance ~ name, paired = TRUE)
                                 #effsize <- wilcox_effsize(data = tmp, formula = spc_abundance ~ name, paired = TRUE)
                                 diff_median_relab <- tmp %>% group_by(name) %>% 
                                     summarise(median_relab = median(spc_abundance)) %>% 
                                     pull(median_relab) %>% diff()
                                 return(tibble(spc = x, 
                                               comparison = comparison, 
                                               wilcox_p = test$p, 
                                               diff_median_relab = diff_median_relab))
                                 
                             }) %>% do.call(., what= rbind)
                             res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                             return(res_test)
                         }) %>% do.call(., what= rbind)

```

## main visualization

```{r echo=FALSE}


tmp <- res_microbiota %>% filter(comparison == "pre_post_FMT") %>% 
    mutate(restored = ifelse(diff_median_relab < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.05, restored, "NS")) %>% 
    select(spc, restored)

tmp2 <- res_microbiota %>% filter(comparison == "post_donor") %>% 
    mutate(type_restoration = ifelse(diff_median_relab < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.05, type_restoration, "normal")) %>%
    select(spc, type_restoration)

tmp <- full_join(tmp, tmp2) %>% mutate(FMT_effect = paste(restored, type_restoration)) %>% 
    mutate(FMT_effect = ifelse(grepl(FMT_effect, pattern = "NS"), "NS", FMT_effect ))
# save in tsv
#write_tsv(tmp, "../data/enrich_test_res_microbiome.tsv", col_names = T )


plot_dt <- mean_relab %>% left_join(tmp %>% filter(FMT_effect != "NS")) %>% na.omit() %>% 
    mutate(spc = str_replace(spc, pattern = "\\|t__", replacement =" ")) %>%
    mutate(spc = str_replace(spc, pattern = "_", replacement =" "))

ggplot(data = plot_dt, 
       aes(x = med_relab, y = reorder(spc, med_relab))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("Average relative abundance (log10)") +
    facet_grid(rows = vars(FMT_effect), 
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank(), 
          axis.text.y = element_text(face = "italic")) + 
    coord_equal(ratio = 0.2)
 

#Amount of each enrichment
tmp %>% group_by(FMT_effect) %>% summarise(nb = n()) %>% 
    mutate(FMT_effect = factor(FMT_effect, levels = names(colormap_enrich)[names(colormap_enrich) %in% FMT_effect])) %>%
     ggplot(data = .,
            aes(fill = FMT_effect, y = nb, x = "FMT effect on SGB")) +
    geom_bar(stat = "identity", position = "fill", 
             col = "black") + 
    geom_text(aes(label = nb), size = 4, hjust = 0.5, position = position_fill(vjust = 0.5)) +
    scale_fill_manual(values = colormap_enrich) + 
    theme_minimal() + xlab("") + ylab("Proportion of SGB") +
    theme(panel.grid.major.x = element_blank(),
          legend.position = "bottom", legend.direction = "vertical") + 
    coord_equal(ratio = 3)

```



## Save results

```{r}
all_measure_mat <- data.frame(cbind(all_measure_mat, t(mpa4_table_spc[unique(plot_dt$spc), ])))
all_measure_type <- all_measure_type %>% rbind(tibble(measure = colnames(all_measure_mat), type = "species_relab"))
write_lines(x = colnames(all_measure_mat), file = "../data/FMT_affected_species.txt")
```


# HUMAnN 3 - Pathways

29/01/2026 - Not used at the moment in our analysis

```{r}
#humann3_dt <- read_tsv(file = "../data/HUMAnN3_pathabundance.tsv", col_names = T, show_col_types = F)
humann3_dt <- read_tsv(file = "../data/HUMAnN3_pathabundance_relab.tsv", col_names = T, show_col_types = F)
humann3_pathway_table <- data.frame(humann3_dt %>% select(-pathway), row.names = humann3_dt %>% pull(pathway), check.names = F )
#pathways_relab <- apply(humann3_pathway_table, 2, function(x) 100*x/sum(x))
pathways_relab <- as.matrix(humann3_pathway_table)
```


## Test

```{r}

pathway_mean_relab <- lapply(rownames(pathways_relab), function(x){
    tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics") %>% distinct(name, sample) %>% filter( sample %in% colnames(pathways_relab)) %>% data.frame()
    tmp$pathway_abundance <- pathways_relab[x, tmp$sample]
    res <- tmp %>% group_by(name) %>% 
        summarise(med_relab = mean(pathway_abundance)) %>%
        mutate(pathway = x)
    return(res)
}) %>% do.call(., what= rbind)

pathway_keep <- apply(pathways_relab, 1, function(x) sum(x > 0)/ncol(pathways_relab) >= 0.5)
#pathway_keep <- apply(pathways_relab, 1, var) %>% sort() %>% head(n = nrow(pathways_relab) * 0.7) %>% names()
pathways_relab_light <- pathways_relab[pathway_keep , ]


res_pathway <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             #grp_torm = "donor"
                             tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")  %>% 
                                 filter(name != comparisons_grp_rm[comparison], sample %in% colnames(pathways_relab)) 
                             tmp <- tmp %>% group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                             pathway_dt <- pathways_relab_light[, tmp$sample]
                           
                             res_test <- lapply(rownames(pathway_dt), function(x){
                                 tmp$pathway_abundance <- pathway_dt[x, ]
                                 if(var(tmp$pathway_abundance) == 0){ return(NULL)}
                                 test <- wilcox_test(data = tmp, formula = pathway_abundance ~ name, paired = TRUE)
                                 diff_median_relab <- tmp %>% group_by(name) %>% 
                                     summarise(median_relab = median(pathway_abundance)) %>% 
                                     pull(median_relab) %>% diff()
                                 return(tibble(pathway = x, 
                                               comparison = comparison, 
                                               wilcox_p = test$p, 
                                               diff_median_relab = diff_median_relab))
                                 
                             }) %>% do.call(., what= rbind)
                             res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                             return(res_test)
                         }) %>% do.call(., what= rbind)

```


```{r echo=FALSE}

tmp <- res_pathway %>% filter(comparison == "pre_post_FMT") %>% 
    mutate(restored = ifelse(diff_median_relab < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.01, restored, "NS")) %>% 
    select(pathway, restored)

tmp2 <- res_pathway %>% filter(comparison == "post_donor") %>% 
    mutate(type_restoration = ifelse(diff_median_relab < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.01, type_restoration, "normal")) %>%
    select(pathway, type_restoration)

tmp <- full_join(tmp, tmp2) %>% mutate(FMT_effect = paste(restored, type_restoration)) 

plot_dt <- pathway_mean_relab %>% left_join(tmp) %>% na.omit() %>%
    mutate(FMT_effect = ifelse(grepl(FMT_effect, pattern = "NS"), "NS", FMT_effect ))


ggplot(data = plot_dt %>% filter(FMT_effect != "NS"), 
       aes(x = med_relab, y = reorder(pathway, med_relab))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("Pathway median relative abundance (log10)") +
    facet_grid(rows = vars(FMT_effect), 
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank(), 
          axis.text.y = element_text(size = 7)) +
    coord_equal(ratio = 0.2)


#Amount of each enrichment
plot_dt %>% group_by(FMT_effect) %>% summarise(nb = n()) %>% 
    mutate(FMT_effect = factor(FMT_effect, levels = names(colormap_enrich)[names(colormap_enrich) %in% FMT_effect])) %>%
     ggplot(data = .,
            aes(fill = FMT_effect, y = nb, x = "FMT effect on bacterial pathways")) +
    geom_bar(stat = "identity", position = "fill", 
             col = "black") + 
    geom_text(aes(label = nb), size = 3, hjust = 0.5, position = position_fill(vjust = 0.5)) +
    scale_fill_manual(values = colormap_enrich) + 
    theme_minimal() + xlab("") + ylab("") +
    theme(panel.grid.major.x = element_blank(),
          legend.position = "right", legend.direction = "vertical") +
        coord_equal(ratio = 3)



```


```{r}
to_add <- data.frame(t(pathways_relab_light[unique(plot_dt$pathway), ]))
all_measure_mat <- cbind(all_measure_mat, to_add)

all_measure_type <- all_measure_type %>% rbind(tibble(measure = colnames(to_add), type = "pathway_relab"))
write_lines(x = colnames(to_add), file = "../data/FMT_affected_pathways.txt")

```



# Metabolites - Targeted

Careful Cosmic_9 missing from metabo data feces

## DATA

```{r}
#metabolomic_dt <- read_tsv(file = "../data/all_metabo_dt.tsv", col_names = T, show_col_types = F) 
#    left_join(pairwise_FMT_dt %>% distinct(FMT_group, sample, name)) %>% as.data.frame()

# Use data with limit of detection value 
metabolomic_dt_limits <- read_tsv(file = "../data/all_metabo_dt_limit_detection.tsv", col_names = T, show_col_types = F) 
metabo_type <- read_tsv(file = "../data/metabolites_types.tsv", col_names = T, show_col_types = F)
```


```{r}

res_metabo <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             #print(comparison)
                             lapply(c("feces", "plasma"), function(SITE){
                                 #print(SITE)
                                 metabo_dt <- metabolomic_dt_limits %>%
                                     filter(bodysite == SITE)  %>% select(-bodysite)
                                 metabo_dt <- t(data.frame(metabo_dt, row.names = "sample"))
                                 
                                 if(SITE == "feces"){
                                     tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics") 
                                 } else {
                                      tmp <- pairwise_FMT_dt %>% filter(data == "blood_scenith")  
                                 }
                                 
                                 tmp <- tmp %>% filter(name != comparisons_grp_rm[comparison]) %>% 
                                     filter(sample %in% colnames(metabo_dt)) %>%
                                     group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                                 metabo_dt <- metabo_dt[, tmp$sample] %>% na.omit()
                                 
                                 res_test <- lapply(rownames(metabo_dt), function(x){
                                     #print(x)
                                     tmp$concentration <- metabo_dt[x, ]
                                     if(var(tmp$concentration) == 0){ return(NULL)}
                                     test <- wilcox_test(data = tmp, formula = concentration ~ name, paired = TRUE)
                                     diff_mean_conc <- tmp %>% group_by(name) %>% 
                                         summarise(mean_conc = median(concentration)) %>% 
                                         pull(mean_conc) %>% diff()
                                     return(tibble(metabolites = x, 
                                                   comparison = comparison, 
                                                   wilcox_p = test$p, 
                                                   diff_mean_conc = diff_mean_conc, 
                                                   bodysite = SITE))
                                     
                                 }) %>% do.call(., what= rbind)
                                 res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                                 return(res_test)
                             }) %>% do.call(., what = rbind)
                         }) %>% do.call(., what = rbind)


mean_conc <-  left_join(pairwise_FMT_dt %>% select(-data, -FMT_group), metabolomic_dt_limits) %>% distinct()
mean_conc <- mean_conc %>% pivot_longer(-c(name, sample, bodysite), names_to = "metabolites", values_to = "concentration")
mean_conc <- mean_conc %>% group_by(name, bodysite, metabolites) %>% summarise(mean_conc = mean(concentration)) 
    
```


main visualization

```{r echo=FALSE}

tmp <- res_metabo %>% filter(comparison == "pre_post_FMT") %>% 
    group_by(bodysite) %>%
    mutate(restored = ifelse(diff_mean_conc < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.05, restored, "NS")) %>% 
    select(metabolites, restored)


tmp2 <- res_metabo %>% filter(comparison == "post_donor") %>% 
    group_by(bodysite) %>%
    mutate(type_restoration = ifelse(diff_mean_conc < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.05, type_restoration, "normal")) %>%
    select(metabolites, type_restoration)

    
all_res <- full_join(tmp, tmp2) %>% 
    mutate(FMT_effect = paste(restored, type_restoration)) %>% 
    mutate(FMT_effect = ifelse(grepl(FMT_effect, pattern = "NS"), "NS", FMT_effect )) 
plot_dt <- mean_conc %>% mutate(metabolites = str_replace_all(string = metabolites, pattern = "-| ", ".")) %>% 
    left_join(all_res) %>% na.omit() %>% 
    left_join(metabo_type %>% mutate(metabolites = str_replace_all(metabolite, "( |-)", ".")))

plot_dt %>% filter(FMT_effect != "NS") %>% arrange(bodysite, type) %>%
    mutate(metabolites = str_replace_all(string = metabolites, pattern = "\\.", " ")) %>%
ggplot(data = ., 
       aes(x = mean_conc, y = reorder(metabolites, mean_conc))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("Median concentration (log10)") +
    facet_grid(rows = vars(FMT_effect), cols = vars(bodysite),
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank()) + 
    coord_equal(ratio = 0.2)


#Amount of each enrichment by metabolite type
res <- all_res %>% mutate(metabolites = str_replace(metabolites, "^X([0-9])", "\\1")) 
res <- res %>% left_join(metabo_type %>% mutate(metabolites = str_replace_all(metabolite, "( |-)", ".")))

res %>% group_by(type, bodysite) %>% mutate(nb_type = n()) %>% 
    group_by( bodysite, type, FMT_effect, nb_type) %>% summarise(nb = n()) %>% 
    mutate(FMT_effect = factor(FMT_effect, levels = names(colormap_enrich))) %>%
    ggplot(data = .,
           aes(fill = FMT_effect, y = nb, x = type)) +
    geom_bar(stat = "identity", position = "fill", 
             col = "black") + 
    geom_text(aes(label = nb), size = 3, hjust = 0.5, position = position_fill(vjust = 0.5)) +
    scale_fill_manual(values = colormap_enrich) + 
    theme_minimal() + xlab("") + ylab("") +
    facet_grid(rows = vars(bodysite)) +
    theme(panel.grid.major.x = element_blank(),
          legend.position = "right", legend.direction = "vertical") + 
    coord_equal(ratio = 3)


```



## Save data

```{r}
targ_metabo_shift <- plot_dt %>% ungroup() %>% filter(FMT_effect != "NS") %>%
    mutate(type = paste("targeted_metabo", bodysite, sep="__"),
           measure = paste(type, metabolites, sep="__")) %>%
    distinct(measure, type, metabolites) 

all_measure_type <- rbind(all_measure_type, targ_metabo_shift %>% select(measure, type))



a <- metabolomic_dt_limits %>% filter(bodysite == "feces") %>% select(-bodysite) %>% data.frame() 
rownames(a) <- a$sample
a <- a[, unique(targ_metabo_shift %>% filter(type == "targeted_metabo__feces") %>% pull(metabolites))]
colnames(a) <- paste0("targeted_metabo__feces__", colnames(a))
all_measure_mat = merge(all_measure_mat, a, by=0, all=TRUE ) 
rownames(all_measure_mat) <- all_measure_mat$Row.names
all_measure_mat <- all_measure_mat %>% select(-Row.names)



b <- metabolomic_dt_limits %>% filter(bodysite == "plasma") %>% select(-bodysite) %>% data.frame() 
rownames(b) <- b$sample
b <- b[, unique(targ_metabo_shift %>% filter(type == "targeted_metabo__plasma") %>% pull(metabolites))]
colnames(b) <- paste0("targeted_metabo__plasma__", colnames(b))
all_measure_mat = merge(all_measure_mat, b, by=0, all=TRUE ) 
rownames(all_measure_mat) <- all_measure_mat$Row.names
all_measure_mat <- all_measure_mat %>% select(-Row.names)



write_lines(c(colnames(a), colnames(b)), file = "../data/FMT_affected_targeted_metabo.txt")

```


# Metabolites - Untargeted

## DATA

```{r}

feces_sample_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN FECES).XLSX", sheet = 3) %>% filter(!is.na(GROUP_ID))
feces_chemical_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN FECES).XLSX", sheet = 2)
feces_log_data <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN FECES).XLSX", sheet = 7)
feces_log_data <- data.frame(feces_log_data %>% select(-PARENT_SAMPLE_NAME), row.names = feces_log_data$PARENT_SAMPLE_NAME)

# Plasma data
plasma_sample_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN PLASMA).XLSX", sheet = 3) %>% filter(!is.na(GROUP_ID))
plasma_chemical_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN PLASMA).XLSX", sheet = 2)
plasma_log_data <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN PLASMA).XLSX", sheet = 7)
plasma_log_data <- data.frame(plasma_log_data %>% select(-PARENT_SAMPLE_NAME), row.names = plasma_log_data$PARENT_SAMPLE_NAME)

```

## Test

```{r}
comparisons_grp_rm2 <- c("pre_post_FMT" = "Healthy", "pre_donor" = "Post_FMT", "post_donor" = "Pre_FMT")

res_metabo_untarg <- lapply(names(comparisons_grp_rm2), 
                         function(comparison){
                             #print(comparison)
                             lapply(c("feces", "plasma"), function(SITE){
                                 #print(SITE)
                                 if(SITE == "feces"){
                                     metabo_measure <- feces_log_data                                         
                                     sample_dt <- feces_sample_metadata
                                     chem_md <- feces_chemical_metadata
                                     fmt_grp_dt <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")
                                 } else {
                                     metabo_measure <- plasma_log_data
                                     sample_dt <- plasma_sample_metadata
                                     chem_md <- plasma_chemical_metadata
                                     fmt_grp_dt <- pairwise_FMT_dt %>% filter(data == "blood_scenith")
                                 }
                                 
                                 sample_dt <- sample_dt %>% dplyr::filter(GROUP_ID != comparisons_grp_rm2[comparison]) %>% 
                                     left_join(fmt_grp_dt %>% select(sample_id = sample, FMT_group)) %>%
                                     group_by(FMT_group) %>% dplyr::filter(n() == 2) %>% ungroup()
                                 metabo_measure <- metabo_measure[sample_dt$PARENT_SAMPLE_NAME, ] %>% na.omit()
                                 
                                 res_test <- lapply(colnames(metabo_measure), function(x){
                                     #print(x)
                                     sample_dt$concentration <- metabo_measure[, x]
                                     if(var(sample_dt$concentration) == 0){ return(NULL)}
                                     test <- wilcox_test(data = sample_dt, formula = concentration ~ GROUP_ID, paired = TRUE)
                                     diff_mean_conc <- sample_dt %>% group_by(GROUP_ID) %>% 
                                         summarise(mean_conc = median(concentration)) %>% 
                                         pull(mean_conc) %>% diff()
                                     return(tibble(metabolites = x, 
                                                   comparison = comparison, 
                                                   wilcox_p = test$p, 
                                                   diff_mean_conc = diff_mean_conc, 
                                                   bodysite = SITE))
                                 }) %>% do.call(., what= rbind)
                                 
                                 res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                                 res_test <- res_test %>% left_join(chem_md %>% 
                                                                        mutate(CHEM_ID = paste0("X", CHEM_ID)) %>% 
                                                                        select(metabolites = CHEM_ID, CHEMICAL_NAME, SUPER_PATHWAY, SUB_PATHWAY) )
                                 
                                 return(res_test)
                             }) %>% do.call(., what = rbind)
                         }) %>% do.call(., what = rbind)


```


```{r}

tmp <- res_metabo_untarg %>% filter(comparison == "pre_post_FMT") %>% 
    group_by(bodysite) %>%
    mutate(restored = ifelse(diff_mean_conc < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.05, restored, "NS")) %>% 
    select(metabolites, restored)


tmp2 <- res_metabo_untarg %>% filter(comparison == "post_donor") %>% 
    group_by(bodysite) %>%
    mutate(type_restoration = ifelse(diff_mean_conc < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.05, type_restoration, "normal")) %>%
    select(metabolites, type_restoration)

    
untarg_metabolites_shift <- full_join(tmp, tmp2) %>% mutate(FMT_effect = paste(restored, type_restoration)) %>%
    mutate(FMT_effect = ifelse(grepl(FMT_effect, pattern = "NS"), "NS", FMT_effect)) %>%
    mutate(metabolites = str_remove(metabolites, "X")) %>% 
    left_join(plasma_chemical_metadata %>% mutate(metabolites = as.character(CHEM_ID)) %>% 
                  select(metabolites, chem_name = PLOT_NAME, SUPER_PATHWAY, SUB_PATHWAY)) %>%
    left_join(feces_chemical_metadata %>% mutate(metabolites = as.character(CHEM_ID)) %>% 
                  select(metabolites, chem_name = PLOT_NAME, SUPER_PATHWAY, SUB_PATHWAY)) %>%
    na.omit() %>%
    mutate(chem_name = paste0(bodysite, "_untarg_", chem_name))

```


## Visualize

plasma

```{r}

mapping_names_untarg_plasma <- plasma_sample_metadata$sample_id
names(mapping_names_untarg_plasma) <- plasma_sample_metadata$PARENT_SAMPLE_NAME
plasma_keep <- plasma_log_data
rownames(plasma_keep) <- mapping_names_untarg_plasma[rownames(plasma_keep)]
metabo_plasma_keep <- untarg_metabolites_shift %>% filter(bodysite == "plasma", FMT_effect != "NS") %>% pull(metabolites)
metabo_plasma_keep <- paste0("X", metabo_plasma_keep)
plasma_keep <- plasma_keep[, metabo_plasma_keep]

x <- plasma_chemical_metadata %>% 
    mutate(metabolites =  paste0("X", CHEM_ID)) %>% select(metabolites, chem_name = PLOT_NAME) %>% 
    filter(metabolites %in% metabo_plasma_keep) %>%
    mutate(chem_name = paste0("plasma_untarg_", chem_name))
mapp_chem <- x$chem_name
names(mapp_chem) <- x$metabolites
colnames(plasma_keep) <- mapp_chem[colnames(plasma_keep)]

plasma_keep$sample_id <- rownames(plasma_keep)

plasma_keep <- plasma_keep %>% pivot_longer(-sample_id, values_to = "untarg_metabo_measure", names_to = "chem_name") %>% 
    mutate(bodysite = "plasma") %>%
    left_join(plasma_sample_metadata %>% select(sample_id, GROUP_ID)) %>% 
    left_join(untarg_metabolites_shift )


plot_dt <- plasma_keep %>% group_by(bodysite, chem_name, GROUP_ID, FMT_effect) %>% 
    summarise(median_value = median(untarg_metabo_measure))

plot_dt %>% mutate(chem_name = str_remove(string = chem_name, "plasma_untarg_")) %>% 
ggplot(data = ., 
       aes(x = median_value, y = reorder(chem_name, median_value))) +
    geom_point(aes(shape = GROUP_ID, fill = GROUP_ID), size = 2) + 
    scale_shape_manual(values = c("Post_FMT" = 21, "Healthy" = 22,  "Pre_FMT" = 23)) +
    scale_fill_manual(values = c("Post_FMT" = "#1f78b4", "Healthy" = "#b2df8a",  "Pre_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("Median value (log10)") +
    facet_grid(rows = vars(FMT_effect), cols = vars(bodysite),
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank()) + coord_equal(ratio = 0.2)


```

fecal

```{r}

mapping_names_untarg_feces <- feces_sample_metadata$sample_id
names(mapping_names_untarg_feces) <- feces_sample_metadata$PARENT_SAMPLE_NAME
feces_keep <- feces_log_data
rownames(feces_keep) <- mapping_names_untarg_feces[rownames(feces_keep)]
metabo_feces_keep <- untarg_metabolites_shift %>% filter(bodysite == "feces", FMT_effect != "NS") %>% pull(metabolites)
metabo_feces_keep <- paste0("X", metabo_feces_keep)
feces_keep <- feces_keep[, metabo_feces_keep]

x <- feces_chemical_metadata %>% 
    mutate(metabolites =  paste0("X", CHEM_ID)) %>% select(metabolites, chem_name = PLOT_NAME) %>% 
    filter(metabolites %in% metabo_feces_keep) %>%
    mutate(chem_name = paste0("feces_untarg_", chem_name))
mapp_chem <- x$chem_name
names(mapp_chem) <- x$metabolites
colnames(feces_keep) <- mapp_chem[colnames(feces_keep)]

feces_keep$sample_id <- rownames(feces_keep)

feces_keep <- feces_keep %>% pivot_longer(-sample_id, values_to = "untarg_metabo_measure", names_to = "chem_name") %>% 
    mutate(bodysite = "feces") %>%
    left_join(feces_sample_metadata %>% select(sample_id, GROUP_ID)) %>% 
    left_join(untarg_metabolites_shift )


plot_dt <- feces_keep %>% group_by(bodysite, chem_name, GROUP_ID, FMT_effect) %>% 
    summarise(median_value = median(untarg_metabo_measure))

plot_dt %>% mutate(chem_name = str_remove(string = chem_name, "feces_untarg_")) %>% 
ggplot(data = ., 
       aes(x = median_value, y = reorder(chem_name, median_value))) +
    geom_point(aes(shape = GROUP_ID, fill = GROUP_ID), size = 2) + 
    scale_shape_manual(values = c("Post_FMT" = 21, "Healthy" = 22,  "Pre_FMT" = 23)) +
    scale_fill_manual(values = c("Post_FMT" = "#1f78b4", "Healthy" = "#b2df8a",  "Pre_FMT" = "#a6cee3")) +
    theme_bw() +
    ylab("") + xlab("Median value") +
    facet_grid(rows = vars(FMT_effect), cols = vars(bodysite),
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank()) + coord_equal(ratio = 0.5)

```


proportion shifting
not looking so good so maybe just keep numbers in case
maybe try by categories of chemical ?

```{r}

untarg_metabolites_shift %>% group_by(bodysite, SUPER_PATHWAY) %>% mutate(tot_nb = n()) %>% 
    group_by(bodysite, FMT_effect, SUPER_PATHWAY, tot_nb) %>% summarize(nb = n()) %>%
    filter(SUPER_PATHWAY != "Partially Characterized Molecules") %>%
    
    ggplot(data = .,
           aes(fill = FMT_effect, y = nb, x = SUPER_PATHWAY)) +
    geom_bar(stat = "identity", position = "fill", 
             col = "black") + 
    geom_text(aes(label = nb), size = 3, hjust = 0.5, position = position_fill(vjust = 0.5)) +
    scale_fill_manual(values = colormap_enrich) + 
    facet_grid(rows =vars(bodysite)) +
    theme_minimal() + xlab("") + ylab("") +
    theme(panel.grid.major.x = element_blank(),
          legend.position = "right", legend.direction = "vertical") + 
    coord_equal(ratio = 3)

```




## add to measure matrix and measure list


```{r}
# mapping between metabolon experiment ID and ours
mapping_names_untarg_plasma <- plasma_sample_metadata$sample_id
names(mapping_names_untarg_plasma) <- plasma_sample_metadata$PARENT_SAMPLE_NAME

# build matrix of only plasma metabo with shift
plasma_keep <- plasma_log_data
rownames(plasma_keep) <- mapping_names_untarg_plasma[rownames(plasma_keep)]
metabo_plasma_keep <- untarg_metabolites_shift %>% filter(bodysite == "plasma", FMT_effect != "NS") %>% pull(metabolites)
metabo_plasma_keep <- paste0("X", metabo_plasma_keep)
plasma_keep <- plasma_keep[, metabo_plasma_keep]

# mapping metabolites names
x <- plasma_chemical_metadata %>% 
    mutate(metabolites =  paste0("X", CHEM_ID)) %>% select(metabolites, chem_name = PLOT_NAME) %>% 
    filter(metabolites %in% metabo_plasma_keep) %>%
    mutate(chem_name = paste0("plasma_untarg_", chem_name))
mapp_chem <- x$chem_name
names(mapp_chem) <- x$metabolites
colnames(plasma_keep) <- mapp_chem[colnames(plasma_keep)]


# merge with other data
plasma_keep[setdiff(rownames(all_measure_mat), rownames(plasma_keep)), ] <- NA
all_measure_mat[setdiff(rownames(plasma_keep), rownames(all_measure_mat)), ] <- NA
all_measure_mat = merge(all_measure_mat, plasma_keep, by=0, all=TRUE ) 
rownames(all_measure_mat) <- all_measure_mat$Row.names
all_measure_mat <- all_measure_mat %>% select(-Row.names)


all_measure_type <- all_measure_type %>% rbind(tibble(measure = colnames(plasma_keep) , type = "plasma_untarg_metabo"))

# feces ----------------------------------------------------------------------- 
mapping_names_untarg_feces <- feces_sample_metadata$sample_id
names(mapping_names_untarg_feces) <- feces_sample_metadata$PARENT_SAMPLE_NAME

feces_keep <- feces_log_data
rownames(feces_keep) <- mapping_names_untarg_feces[rownames(feces_keep)]
metabo_feces_keep <- untarg_metabolites_shift %>% filter(bodysite == "feces", FMT_effect != "NS") %>% pull(metabolites)
metabo_feces_keep <- paste0("X", metabo_feces_keep)
feces_keep <- feces_keep[, metabo_feces_keep]

x <- feces_chemical_metadata %>% 
    mutate(metabolites =  paste0("X", CHEM_ID)) %>% select(metabolites, chem_name = PLOT_NAME) %>% 
    filter(metabolites %in% metabo_feces_keep) %>%
    mutate(chem_name = paste0("feces_untarg_", chem_name))
mapp_chem <- x$chem_name
names(mapp_chem) <- x$metabolites
colnames(feces_keep) <- mapp_chem[colnames(feces_keep)]

feces_keep[setdiff(rownames(all_measure_mat), rownames(feces_keep)), ] <- NA
all_measure_mat = merge(all_measure_mat, feces_keep, by=0, all=TRUE ) 
rownames(all_measure_mat) <- all_measure_mat$Row.names
all_measure_mat <- all_measure_mat %>% select(-Row.names)
all_measure_type <- all_measure_type %>% rbind(tibble(measure = colnames(feces_keep), type = "feces_untarg_metabo"))


write_lines(x = c(colnames(feces_keep), colnames(plasma_keep)), file = "../data/FMT_affected_untargeted_metabo.txt")

```


save updated results 

```{r}

saveRDS(object = all_measure_mat, file = "../data/all_measure_signif.rds")
write_tsv(all_measure_type, file = "../data/all_signif_measure_type.tsv", col_names = T)

```





## Focus on AA levels

```{r}

aa_list <- c("histidine", "isoleucine","leucine","lysine","methionine",
             "phenylalanine","threonine","tryptophan","valine","arginine",
             "cysteine","glutamine","tyrosine","glycine", "proline", "serine")

SITE = "feces"

if(SITE == "feces"){
    metabo_measure <- feces_log_data                                         
    sample_dt <- feces_sample_metadata
    chem_md <- feces_chemical_metadata
    fmt_grp_dt <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")
} else {
    metabo_measure <- plasma_log_data
    sample_dt <- plasma_sample_metadata
    chem_md <- plasma_chemical_metadata
    fmt_grp_dt <- pairwise_FMT_dt %>% filter(data == "blood_scenith")
}

sample_dt <- sample_dt %>%  left_join(fmt_grp_dt %>% select(sample_id = sample, FMT_group, name)) 
metabo_measure <- metabo_measure[sample_dt$PARENT_SAMPLE_NAME, ] %>% na.omit()

#TARGET_ID <- chem_md %>% filter(CHEMICAL_NAME %in% c("2-piperidinone", "N-methylproline", "proline")) %>% pull(CHEM_ID)
TARGET_ID <- chem_md %>% filter(CHEMICAL_NAME %in% aa_list) %>% pull(CHEM_ID)

tmp <- metabo_measure[, paste0("X", TARGET_ID)]
# colnames(tmp) <-  chem_md %>% filter(CHEMICAL_NAME %in% c("2-piperidinone", "N-methylproline", "proline")) %>% pull(CHEMICAL_NAME)
colnames(tmp) <-  chem_md %>% filter(CHEMICAL_NAME %in% aa_list) %>% pull(CHEMICAL_NAME)
tmp$PARENT_SAMPLE_NAME <- rownames(tmp)


plot_dt <- tmp %>% left_join(sample_dt %>% select(PARENT_SAMPLE_NAME, name)) %>% na.omit() %>%
    pivot_longer(-c(PARENT_SAMPLE_NAME, name), values_to = "chem_concentration", names_to = "chem_name")

plot_dt$name <- factor(plot_dt$name, levels = c("before_FMT",  "donor", "after_FMT"))

ggplot(data = plot_dt,
       aes(x = name, y = chem_concentration, color = name)) + 
    geom_boxplot() + 
    ggpubr::stat_pwc(tip.length = 0, p.adjust.method = "BH", p.adjust.by = "panel", hide.ns = T) +
    geom_jitter(height = 0)  +
    theme_bw() + ylab("Normalized concentration") + xlab("") +
    scale_color_manual(values = colormap_group)  +
    facet_wrap(~chem_name, ncol = 8) +
    ggtitle(SITE) + 
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 90),
          strip.background = element_blank())

 
```


# SCENITH

Pairwise test before after

```{r}

scenith_dt <- read_tsv(file = "../data/scenith_data.tsv", col_names = T, show_col_types = F)

test_dt <- scenith_dt %>% 
    pivot_longer(-c(type_measure, group, sample), names_to = "cell_type", values_to = "SCENITH_measure") %>% 
    filter(group != "HD") %>%
    left_join(pairwise_FMT_dt %>% filter(data == "blood_scenith", name != "donor") %>% select(FMT_group, sample)) %>%
    mutate(cell_measure = paste(cell_type, type_measure, sep="__"))

res_test <- lapply(unique(test_dt$cell_measure), function(x){
    
    tmp <- test_dt %>% filter(cell_measure == x) %>%
        group_by(FMT_group) %>% filter(n() ==  2) %>% arrange(FMT_group) %>% ungroup()
    
    test <- wilcox_test(data = tmp, formula = SCENITH_measure ~ group, paired = TRUE)
    effsize <- wilcox_effsize(data = tmp, formula = SCENITH_measure ~ group, paired = TRUE)
    
    return(tibble(cell_measure = x, 
                  wilcox_p = test$p, 
                  wilcox_effsize = effsize$effsize))
}) %>% do.call(., what = rbind) %>% mutate(wilcox_Padj = p.adjust(wilcox_p, method = "BH"))


res_test_signif <- res_test %>% filter(wilcox_Padj < 0.05)

res_test_signif <- res_test_signif %>% 
    mutate(type_measure = str_remove_all(cell_measure, pattern = ".*__" ), 
           cell_type = str_remove_all(cell_measure, pattern = "__.*")) %>% 
    select(cell_type, type_measure, wilcox_Padj)

```


```{r}

cell_measure_signif <- res_test %>% filter(wilcox_Padj < 0.05) %>% pull(cell_measure)


export_dt <- scenith_dt %>% 
    pivot_longer(-c(type_measure, group, sample), names_to = "cell_type", values_to = "SCENITH_measure") %>% 
    left_join(pairwise_FMT_dt %>% filter(data == "blood_scenith", name != "donor") %>% select(FMT_group, sample)) %>%
    mutate(cell_measure = paste(cell_type, type_measure, sep="__"))

scenith_dt_signif <- export_dt %>% filter(cell_measure %in% cell_measure_signif) %>% 
    select(cell_measure, SCENITH_measure, sample) %>% 
    pivot_wider(names_from = cell_measure, values_from = SCENITH_measure) 
scenith_dt_signif <- data.frame(scenith_dt_signif %>% select(-sample), row.names = scenith_dt_signif$sample)

all_measure_mat <- merge(all_measure_mat, scenith_dt_signif, by = "row.names", all = TRUE)
rownames(all_measure_mat) <- all_measure_mat$Row.names
all_measure_mat <- all_measure_mat %>% select(-Row.names)

all_measure_type <- rbind(all_measure_type, tibble(measure = colnames(scenith_dt_signif), type = "SCENITH_measure"))


```


Overall doorbell plot with shift to mean of HD

```{r}
scenith_dt <- read_tsv(file = "../data/scenith_data.tsv", col_names = T, show_col_types = F)

plot_dt <- scenith_dt %>% select(-sample) %>% 
    pivot_longer(-c(type_measure, group), names_to = "cell_type", values_to = "SCENITH_measure") %>%
    group_by(type_measure, group, cell_type) %>% summarise(avg_SCENITH = mean(SCENITH_measure)) %>%
    pivot_wider(id_cols = c(type_measure, cell_type), 
                values_from = avg_SCENITH, 
                names_from = group, names_prefix = "avg_SCENITH_") %>%
    mutate(shift_after_FMT = avg_SCENITH_HD - avg_SCENITH_after_FMT, 
           shift_before_FMT = avg_SCENITH_HD - avg_SCENITH_before_FMT) %>% 
    select(-avg_SCENITH_HD, -avg_SCENITH_after_FMT, -avg_SCENITH_before_FMT) %>%
    pivot_longer(cols = c(shift_after_FMT, shift_before_FMT), 
                 names_to = "before_after", values_to = "shift") 

plot_dt <- plot_dt %>% left_join(res_test_signif) %>% 
    mutate(color = ifelse(is.na(wilcox_Padj), "NS", before_after))


ggplot(data = plot_dt, 
       aes(x = shift, y = reorder(cell_type, shift))) + 
    geom_segment( aes(x=shift, xend=0, y=cell_type, yend=cell_type), color="grey") +
    geom_point(aes(shape = before_after, fill = color), size = 3) + 
    geom_vline(xintercept = 0) + 
    scale_shape_manual(values = c("shift_after_FMT" = 21, "shift_before_FMT" = 23), 
                       labels = c("shift_after_FMT" = "after", "shift_before_FMT" = "before")) +
    scale_fill_manual(values = c("shift_after_FMT" = "#1f78b4",  "shift_before_FMT" = "#a6cee3", 
                                 "NS" = "white"),
                      guide = "none") +
    theme_bw() + 
    guides(shape = guide_legend(title = "Before/After FMT")) +
    ylab("") + xlab("Average SCENITH measurement") +
    facet_grid(cols = vars(type_measure), scale = "free")

```




```{r}

saveRDS(object = all_measure_mat, file = "../data/all_measure_signif.rds")
write_tsv(all_measure_type, file = "../data/all_signif_measure_type.tsv", col_names = T)

```


# Targeted Correlations

do correlation beetween stuff and visualize as graph
build matrix with pairwise correlation of elements (metabo, spc, scenith)

1 element : 
- scenith, 1 type measure, 
- species abundance
- metabo in bodysite

## Correlation (initial screening plan)

Screen all pairwise correlation between all measures

```{r}
all_measure_mat <- readRDS(file = "../data/all_measure_signif.rds")
all_measure_type <- read_tsv(file = "../data/all_signif_measure_type.tsv", col_names = T, show_col_types = F)
```


```{r warning=FALSE}

sample_keep <- pairwise_FMT_dt %>% filter(name != "before_FMT") %>% pull(sample) %>% unique()
data_for_corr_test <- all_measure_mat[sample_keep, all_measure_type %>% filter(type != "pathway_relab") %>% pull(measure) ]

corr_res <- lapply(combn(colnames(data_for_corr_test), m = 2, simplify = F), 
       function(x){
           tmp_dt <- data_for_corr_test[, x] %>% na.omit()
           res <- cor.test(tmp_dt[, 1], tmp_dt[, 2], method = "spearman")
           return(tibble(measure1 = x[1], 
                         measure2 = x[2], 
                         rho = res$estimate, 
                         p = res$p.value))
       }) %>% do.call(., what = rbind)

corr_res_type <- corr_res %>% 
    left_join(all_measure_type %>% select(measure1 = measure, type1 = type)) %>% 
    left_join(all_measure_type %>% select(measure2 = measure, type2 = type)) %>%
    filter(type1 != type2) %>%
    mutate(Padj = p.adjust(p, method = "BH"))
write_tsv(corr_res_type, file = "../data/correlation_results_all.tsv", col_names = T)

#write_tsv(corr_res_type %>% filter(Padj < 0.05), file = "../data/correlation_results_signif.tsv", col_names = T)

```


## Second plan

look first for diff pre post FMT metabo and corr with SCNENITH, 
then look corr these metabo with taxa

```{r}
target_taxa <- read_lines(file = "../data/FMT_affected_species.txt")
target_metabo <- read_lines(file = "../data/FMT_affected_targeted_metabo.txt")
target_metabo_untarg <- read_lines(file = "../data/FMT_affected_untargeted_metabo.txt")

corr_res_type <- read_tsv(file = "../data/correlation_results_all.tsv", col_names = T, show_col_types = F)

# filter first metabo x SCENITH : corr1 
corr_1 <- corr_res_type %>% filter(measure1 %in% c(target_metabo, target_metabo_untarg) | measure2 %in% c(target_metabo, target_metabo_untarg))
corr_1 <- corr_1 %>% filter(type1 == "SCENITH_measure" | type2 == "SCENITH_measure")
# focus monocytes ?
corr_1 <- corr_1 %>% filter(grepl(measure1, pattern = "Mono") | grepl(measure2, pattern = "Mono"))
corr_1$Padj <- p.adjust(corr_1$p, method = "BH") # recompute correction pvalue
write_tsv(corr_1, file = "../data/correlation_1_signif01_metabo_scenith.tsv", col_names = T)



# keep only metabo shifting with FMT, having an impact on SCENITH
metabo_target <- c(corr_1 %>% filter(Padj < 0.05, type1 != "SCENITH_measure") %>% pull(measure1),
                   corr_1 %>% filter(Padj < 0.05, type2 != "SCENITH_measure") %>% pull(measure2))
# filter then metabo x taxa: corr2
corr_2 <- corr_res_type %>% filter(measure1 %in% metabo_target | measure2 %in% metabo_target)
corr_2 <- corr_2 %>% filter(measure1 %in% target_taxa | measure2 %in% target_taxa)
corr_2$Padj <- p.adjust(corr_2$p, method = "BH") # recompute correction pvalue

write_tsv(corr_2, file = "../data/correlation_2_signif01_metabo_taxa.tsv", col_names = T)

#lol = rbind(corr1_signif, corr_2_signif)

```


## Third plan

Corr Metabo x SCENITH with all data
Then corr Metabo x taxa without the before FMT group

```{r}
all_measure_mat <- readRDS(file = "../data/all_measure_signif.rds")
all_measure_type <- read_tsv(file = "../data/all_signif_measure_type.tsv", col_names = T, show_col_types = F)
target_taxa <- read_lines(file = "../data/FMT_affected_species.txt")
target_metabo <- read_lines(file = "../data/FMT_affected_targeted_metabo.txt")
target_metabo_untarg <- read_lines(file = "../data/FMT_affected_untargeted_metabo.txt")


sample_keep <- pairwise_FMT_dt %>% pull(sample) %>% unique()
measure_keep <- all_measure_type %>% 
#    filter(type == "SCENITH_measure" | measure %in% c(target_metabo, target_metabo_untarg)) %>%
    filter(grepl(measure, pattern = "Mono") | measure %in% c(target_metabo, target_metabo_untarg)) %>%

    pull(measure)
data_for_corr_test <- all_measure_mat[sample_keep, measure_keep ]

corr_res1 <- lapply(combn(colnames(data_for_corr_test), m = 2, simplify = F), 
       function(x){
           tmp_dt <- data_for_corr_test[, x] %>% na.omit()
           res <- cor.test(tmp_dt[, 1], tmp_dt[, 2], method = "spearman")
           return(tibble(measure1 = x[1], 
                         measure2 = x[2], 
                         rho = res$estimate, 
                         p = res$p.value))
       }) %>% do.call(., what = rbind)
corr_res1_mono <- corr_res1 %>% 
    left_join(all_measure_type %>% select(measure1 = measure, type1 = type)) %>% 
    left_join(all_measure_type %>% select(measure2 = measure, type2 = type)) %>%
    filter(grepl(measure1, pattern = "Mono") | grepl(measure2, pattern = "Mono")) %>%
    filter(type1 != type2) %>%
    mutate(Padj = p.adjust(p, method = "BH"))

corr_res1_mono <- corr_res1_mono %>% filter(Padj <= 0.05)

```




```{r}
sample_keep <- pairwise_FMT_dt %>%  filter(name != "before_FMT") %>% pull(sample) %>% unique()
target_metabo_pro <- c("plasma_untarg_2-piperidinone", "feces_untarg_2-piperidinone",
                   "feces_untarg_N-methylproline", "plasma_untarg_N-methylproline",
                   "plasma_untarg_proline", "feces_untarg_proline", "feces_untarg_prolylhydroxyproline")
measure_keep <- all_measure_type %>% filter(measure %in% target_taxa  | measure %in% target_metabo_pro) %>% pull(measure)
#measure_keep <- all_measure_type %>% filter(measure %in% target_taxa  | measure %in% c(corr_res1$measure1)) %>% pull(measure)

data_for_corr_test <- all_measure_mat[sample_keep, measure_keep ]

corr_res2 <- lapply(combn(colnames(data_for_corr_test), m = 2, simplify = F), 
       function(x){
           tmp_dt <- data_for_corr_test[, x] %>% na.omit()
           res <- cor.test(tmp_dt[, 1], tmp_dt[, 2], method = "spearman")
           return(tibble(measure1 = x[1], 
                         measure2 = x[2], 
                         rho = res$estimate, 
                         p = res$p.value))
       }) %>% do.call(., what = rbind)
corr_res2 <- corr_res2 %>% 
    left_join(all_measure_type %>% select(measure1 = measure, type1 = type)) %>% 
    left_join(all_measure_type %>% select(measure2 = measure, type2 = type)) %>%
    filter(type1 == "species_relab" | type2 == "species_relab" ) %>%
    filter(type1 != type2) %>%
    mutate(Padj = p.adjust(p, method = "BH"))

corr_res2 %>% filter(Padj <= 0.05)


```


Test taxa with presence associated with pro derivative 


```{r}

target_taxa <- read_lines(file = "../data/FMT_affected_species.txt")

TARGET = "feces_untarg_N-methylproline"

data_for_test <- all_measure_mat[ ,  c(TARGET, target_taxa ) ]

res_taxa_proline <- lapply(target_taxa, function(taxa){
    tmp_dt <- data_for_test[, c(taxa, TARGET)] %>% na.omit()
    colnames(tmp_dt)[2] = "TARGET"
    tmp_dt$sample <- rownames(tmp_dt)
    tmp_dt <- tmp_dt %>% left_join(pairwise_FMT_dt %>% distinct(name, sample))
    tmp_dt$taxa_presence <- ifelse(tmp_dt[,taxa] == 0, "absent", "present")
    res_wilcox <- wilcox_test(data = tmp_dt, TARGET ~ taxa_presence)
    res_wilcox_effsize <- wilcox_effsize(data = tmp_dt, TARGET ~ taxa_presence)
    dt_corr <- tmp_dt %>% filter(taxa_presence == "present")
    res_cor <- cor.test(dt_corr[, "TARGET"], dt_corr[,taxa], 
                        method = "spearman")
    
    return(tibble(taxa = taxa, 
                  wilcox_p = res_wilcox$p, 
                  wilcox_effsize = res_wilcox_effsize$effsize, 
                  rho = res_cor$estimate, 
                  cor_p = res_cor$p.value))
}) %>% do.call(., what = rbind) %>%
    mutate(wilcox_Padj = p.adjust(wilcox_p, "BH"), 
           cor_Padj = p.adjust(cor_p, method = "BH"))
 
res_taxa_proline

```




## Visualize heatmap correlation ALL METABO x SCENITH

```{r}
corr_res_type <- read_tsv(file = "../data/correlation_results_all.tsv", col_names = T, show_col_types = F)
corr_res_type <- read_tsv(file = "../data/correlation_1_signif01_metabo_scenith.tsv", col_names = T, show_col_types = F)

TARG1 = c("feces_untarg_metabo", "plasma_untarg_metabo", "targeted_metabo__plasma", "targeted_metabo__feces")
TARG2 = "SCENITH_measure"
TARG = c(TARG1, TARG2)

dt_scenith_metabo <- corr_res_type %>% filter(type1 %in% TARG & type2 %in% TARG)
dt_scenith_metabo <- dt_scenith_metabo %>% filter(type1 == TARG2 |  type2 == TARG2)

to_inverse <- which(dt_scenith_metabo$type1 == TARG2)
tmp_switch <- dt_scenith_metabo[to_inverse, "measure1"]
dt_scenith_metabo[to_inverse, "measure1"] <- dt_scenith_metabo[to_inverse, "measure2"]
dt_scenith_metabo[to_inverse, "measure2"] <- tmp_switch

tmp_switch <- dt_scenith_metabo[to_inverse, "type1"]
dt_scenith_metabo[to_inverse, "type1"] <- dt_scenith_metabo[to_inverse, "type2"]
dt_scenith_metabo[to_inverse, "type2"] <- tmp_switch

dt_scenith_metabo$measure2 <- str_replace_all(dt_scenith_metabo$measure2, "__", " ")
dt_scenith_metabo <- na.omit(dt_scenith_metabo) 


# keep only measure with at least one corr
measure_signif <- dt_scenith_metabo %>% filter(Padj < 0.05) %>% select(measure1, measure2)
measure_signif <- unique(c(measure_signif$measure1, measure_signif$measure2))
dt_scenith_metabo <- dt_scenith_metabo %>% filter(measure1 %in% measure_signif & measure2 %in% measure_signif)

# prepare Heatmap
type_measure_scenith <- unique(dt_scenith_metabo$measure2) %>% str_remove(pattern = ".* ") 
names(type_measure_scenith) <- unique(dt_scenith_metabo$measure2)

tmp = dt_scenith_metabo %>% distinct(type1, measure1)
type_measure_metabo <- tmp$type1
names(type_measure_metabo) <- tmp$measure1

corr_dt <- dt_scenith_metabo %>% 
    select(measure1, measure2, rho) %>% 
    pivot_wider(names_from = measure2, values_from = rho) %>%
    as.data.frame()
rownames(corr_dt) <- corr_dt$measure1
corr_dt <- corr_dt %>% select(-measure1)
corr_dt[is.na(corr_dt)] <- 0

pval_dt <- dt_scenith_metabo %>% 
    select(measure1, measure2, Padj) %>% 
    pivot_wider(names_from = measure2, values_from = Padj) %>%
    as.data.frame()
rownames(pval_dt) <- pval_dt$measure1
pval_dt <- pval_dt %>% select(-measure1)


col_fun_HM <- colorRamp2(c("blue", "white", "red"),
                             breaks = c(1, 0, -1))
h <- Heatmap(matrix = corr_dt, 
            col = col_fun_HM,
            na_col = "white",
                 
            width = unit(0.5, "cm") * ncol(corr_dt), 
            height = unit(0.5, "cm") * nrow(corr_dt),
            
            name = "SCENITH x Metabolites",
            column_title_gp = gpar(fontsize = 12),
            column_names_rot = 45,
            cluster_rows = T, 
            show_row_dend = F,
            cluster_columns = T, 
            show_column_names = T,
            show_row_names = T, column_title = NULL,
            row_title_side = "right", row_title_rot = 0,
            row_labels = structure(names = rownames(corr_dt), str_remove(rownames(corr_dt), ".*_")),
            
            cluster_column_slices = F,
            show_column_dend = F, 
            clustering_distance_rows = "spearman",
            clustering_distance_columns = "spearman",
            clustering_method_rows = "ward.D2",
            clustering_method_columns = "ward.D2",
            
            row_split = type_measure_metabo, 
            column_split = type_measure_scenith,
            
            border = T,                
            column_gap = unit(2, "mm"), row_gap = unit(2, "mm"),
            row_names_gp = gpar(fontsize = 9),     
            column_names_gp = gpar(fontsize = 9),
            
            rect_gp = gpar(col = "white", lwd = 0.5),
            show_heatmap_legend =T,
            row_names_side = "left", 
            
            cell_fun = function(j, i, x, y, width, height, fill) {
                if(!is.na(pval_dt[i, j] )){
                    if(pval_dt[i, j] < 0.01){
                        color_star <- ifelse(abs(corr_dt[i, j]) > 0.6, "white", "black")
                        grid.text("**", x, y -height/3, gp = gpar(fontsize = 14, col = color_star))
                    } else
                    if(pval_dt[i, j] < 0.05){
                        color_star <- ifelse(abs(corr_dt[i, j]) > 0.6, "white", "black")
                        grid.text("*", x, y -height/3, gp = gpar(fontsize = 14, col = color_star))
                    }
                }
            })

draw(h, heatmap_legend_side = "right", )

```

## Targeted set 1 SCENITH x METABO 

```{r}
dt_scenith_metabo <- corr_res1_mono 

dt_scenith_metabo <- read_tsv(file = "../data/correlation_1_signif01_metabo_scenith.tsv", col_names = T, show_col_types = F)
dt_scenith_metabo$measure2 <- str_replace_all(dt_scenith_metabo$measure2, "__", " ")
dt_scenith_metabo <- na.omit(dt_scenith_metabo) 

# keep only measure with at least one corr
measure_signif <- dt_scenith_metabo %>% filter(Padj < 0.05) %>% select(measure1, measure2)
measure_signif <- unique(c(measure_signif$measure1, measure_signif$measure2))
dt_scenith_metabo <- dt_scenith_metabo %>% filter(measure1 %in% measure_signif & measure2 %in% measure_signif)

type_measure_scenith <- unique(dt_scenith_metabo$measure2) %>% str_remove(pattern = ".* ") 
names(type_measure_scenith) <- unique(dt_scenith_metabo$measure2)

tmp = dt_scenith_metabo %>% distinct(type1, measure1)
type_measure_metabo <- tmp$type1
names(type_measure_metabo) <- tmp$measure1

corr_dt <- dt_scenith_metabo %>% 
    select(measure1, measure2, rho) %>% 
    pivot_wider(names_from = measure1, values_from = rho) %>%
    as.data.frame()
rownames(corr_dt) <- corr_dt$measure2
corr_dt <- corr_dt %>% select(-measure2)
corr_dt[is.na(corr_dt)] <- 0

pval_dt <- dt_scenith_metabo %>% 
    select(measure1, measure2, Padj) %>% 
    pivot_wider(names_from = measure1, values_from = Padj) %>%
    as.data.frame()
rownames(pval_dt) <- pval_dt$measure2
pval_dt <- pval_dt %>% select(-measure2)


col_fun_HM <- colorRamp2(c("blue", "white", "red"),
                             breaks = c(1, 0, -1))
h <- Heatmap(matrix = corr_dt, 
            col = col_fun_HM,
            na_col = "white",
                 
            width = unit(0.35, "cm") * ncol(corr_dt), 
            height = unit(0.35, "cm") * nrow(corr_dt),
            
            name = "SCENITH x Metabolites",
            column_title_gp = gpar(fontsize = 12),
            column_title_side = "top", 
            cluster_rows = T, 
            show_row_dend = F,
            cluster_columns = T, 
            show_column_names = T,
            show_row_names = T, row_title = NULL,
            
            cluster_column_slices = F,
            show_column_dend = F,
            clustering_distance_rows = "spearman",
            clustering_distance_columns = "spearman",
            clustering_method_rows = "ward.D2",
            clustering_method_columns = "ward.D2",
            
            row_split = type_measure_scenith, 
            column_split = type_measure_metabo,
            
            border = T,                
            column_gap = unit(2, "mm"), row_gap = unit(2, "mm"),
            row_names_gp = gpar(fontsize = 9),     
            column_names_gp = gpar(fontsize = 9),
            
            rect_gp = gpar(col = "white", lwd = 0.5),
            show_heatmap_legend =T,
            row_names_side = "left", 
            
            cell_fun = function(j, i, x, y, width, height, fill) {
                if(!is.na(pval_dt[i, j] )){
                    if(pval_dt[i, j] < 0.01){
                        color_star <- ifelse(abs(corr_dt[i, j]) > 0.6, "white", "black")
                        grid.text("**", x, y -height/3, gp = gpar(fontsize = 12, col = color_star))
                    } else
                    if(pval_dt[i, j] < 0.05){
                        color_star <- ifelse(abs(corr_dt[i, j]) > 0.6, "white", "black")
                        grid.text("*", x, y -height/3, gp = gpar(fontsize = 12, col = color_star))
                    }
                }
            })

draw(h, heatmap_legend_side = "right", )

```

## Targeted set 2 SPECIES x Metabo

```{r}

dt_scenith_metabo <- read_tsv(file = "../data/correlation_2_signif01_metabo_taxa.tsv", col_names = T, show_col_types = F)
dt_scenith_metabo$measure2 <- str_replace_all(dt_scenith_metabo$measure2, "__", " ")
dt_scenith_metabo <- na.omit(dt_scenith_metabo) 

# keep only measure with at least one corr
measure_signif <- dt_scenith_metabo %>% filter(Padj < 0.05) %>% select(measure1, measure2)
measure_signif <- unique(c(measure_signif$measure1, measure_signif$measure2))
dt_scenith_metabo <- dt_scenith_metabo %>% filter(measure1 %in% measure_signif & measure2 %in% measure_signif)

tmp = dt_scenith_metabo %>% distinct(type2, measure2)
type_measure_metabo <- tmp$type2
names(type_measure_metabo) <- tmp$measure2

corr_dt <- dt_scenith_metabo %>% 
    select(measure1, measure2, rho) %>% 
    pivot_wider(names_from = measure1, values_from = rho) %>%
    as.data.frame()
rownames(corr_dt) <- corr_dt$measure2
corr_dt <- corr_dt %>% select(-measure2)
corr_dt[is.na(corr_dt)] <- 0

pval_dt <- dt_scenith_metabo %>% 
    select(measure1, measure2, Padj) %>% 
    pivot_wider(names_from = measure1, values_from = Padj) %>%
    as.data.frame()
rownames(pval_dt) <- pval_dt$measure2
pval_dt <- pval_dt %>% select(-measure2)


col_fun_HM <- colorRamp2(c("blue", "white", "red"),
                             breaks = c(1, 0, -1))
h <- Heatmap(matrix = corr_dt, 
            col = col_fun_HM,
            na_col = "white",
                 
            width = unit(0.35, "cm") * ncol(corr_dt), 
            height = unit(0.3, "cm") * nrow(corr_dt),
            
            name = "SCENITH x Metabolites",
            column_title_gp = gpar(fontsize = 12),
            column_title_side = "top", column_names_rot = 45,  
            cluster_rows = T, 
            show_row_dend = F,
            cluster_columns = T, 
            show_column_names = T,
            show_row_names = T, 
            row_title_side = "right", row_title_rot = 0,
            row_labels = structure(names = rownames(corr_dt), str_remove(rownames(corr_dt), ".*_")),

            
            cluster_column_slices = F,
            show_column_dend = F,
            clustering_distance_rows = "spearman",
            clustering_distance_columns = "spearman",
            clustering_method_rows = "ward.D2",
            clustering_method_columns = "ward.D2",
            
            row_split = type_measure_metabo,
            
            border = T,                
            column_gap = unit(2, "mm"), row_gap = unit(2, "mm"),
            row_names_gp = gpar(fontsize = 9),     
            column_names_gp = gpar(fontsize = 9),
            
            rect_gp = gpar(col = "white", lwd = 0.5),
            show_heatmap_legend =T,
            row_names_side = "left", 
            
            cell_fun = function(j, i, x, y, width, height, fill) {
                if(!is.na(pval_dt[i, j] )){
                    if(pval_dt[i, j] < 0.01){
                        color_star <- ifelse(abs(corr_dt[i, j]) > 0.6, "white", "black")
                        grid.text("**", x, y -height/3, gp = gpar(fontsize = 12, col = color_star))
                    } else
                    if(pval_dt[i, j] < 0.05){
                        color_star <- ifelse(abs(corr_dt[i, j]) > 0.6, "white", "black")
                        grid.text("*", x, y -height/3, gp = gpar(fontsize = 12, col = color_star))
                    }
                }
            })

draw(h, heatmap_legend_side = "right", )

```


## Visualize as graph

```{r}
library(igraph)
library(viridis)

lol <- read_tsv(file = "../data/correlation_results_all_with_pathways.tsv", col_names = T, show_col_types = F) %>% filter(Padj < 0.05)
lol$measure1 <- str_remove_all(lol$measure1, pattern = ".*:.")
lol$measure2 <- str_remove_all(lol$measure2, pattern = ".*:.")


scenith_corr_signif <- lol %>% filter(type1 == "SCENITH_measure" | type2 == "SCENITH_measure")
measure_keep <- unique(c(scenith_corr_signif$measure1, scenith_corr_signif$measure2))
lol <- lol %>% filter(measure1 %in% measure_keep | measure2 %in% measure_keep)

# OR
lol <- lol %>% filter(type1 == "SCENITH_measure" | type2 == "SCENITH_measure")


# focus on Monocytes
#lol <- lol %>% filter(grepl(x= measure2, "Mono"))

links <- data.frame(source=lol$measure1,
                    target=lol$measure2,
                    importance=lol$rho )

nodes <- data.frame(rbind(lol %>% select(name = measure1, type = type1), 
                          lol %>% select(name = measure2, type = type2)) %>% 
                        distinct())

g <- graph_from_data_frame(d = links, vertices = nodes , directed = F)
#g <- graph_from_data_frame(d = lol, directed = F)


coul  <- viridis(length(unique(nodes$type))) 
my_color <- coul[as.numeric(as.factor(V(g)$type))]

col_edge <- ifelse(E(g)$importance > 0, "#2c7bb6", "#d7191c")
 
# Make the plot
plot(g, 
     vertex.color=my_color,
     edge.color  = col_edge,
     vertex.size = 3,
     vertex.label.color="black", 
     vertex.label.cex=0.5, 
     layout=layout_with_fr)



pdf(file = "../figures/SCENITH_measure_graph_corr.pdf", height = 15, width = 20)
plot(g, 
     vertex.color=my_color,
     edge.color  = col_edge,
     vertex.size = 3,
     vertex.label.color="black", 
     vertex.label.cex=0.5, 
     layout=layout_with_fr)
dev.off()


```


```{r}

# graph might not be the best visualization 
# TODO scatter


plot_SCENITH_measure_focus <- function(x = "Mono.intermediate__Mitochondrial.dependence", second_connection = FALSE){
    
    dt <- read_tsv(file = "../data/correlation_results_signif.tsv", col_names = T, show_col_types = F) 
    dt$measure1 <- str_remove_all(dt$measure1, pattern = ".*:.")
    dt$measure2 <- str_remove_all(dt$measure2, pattern = ".*:.")
    lol <- dt %>% filter(measure1 == x | measure2 == x)
    
    if(second_connection) {
        first_connection <- unique(c(lol$measure1, lol$measure2))
        lol2 <- dt %>% filter(measure1 %in% first_connection | measure2 %in% first_connection)
        lol <- rbind(lol, lol2)
    }
    
    links <- data.frame(source=lol$measure1,
                        target=lol$measure2,
                        importance=lol$rho )
    nodes <- data.frame(rbind(lol %>% select(name = measure1, type = type1), 
                              lol %>% select(name = measure2, type = type2)) %>% 
                            distinct())
    g <- graph_from_data_frame(d = links, vertices = nodes , directed = F)

    coul  <- viridis(length(unique(nodes$type))) 
    my_color <- coul[as.numeric(as.factor(V(g)$type))]
    
    col_edge <- ifelse(E(g)$importance > 0, "#2c7bb6", "#d7191c")
    
    # Make the plot
    plot(g, 
         vertex.color= my_color,
         edge.color  = col_edge,
         edge.width=abs(E(g)$importance)*2,
         vertex.size = 4,
         vertex.label.color="black", 
         vertex.label.cex=0.8, 
         layout=layout_with_fr)
    
}

plot_SCENITH_measure_focus()

```


```{r}

plot_SCENITH_measure_focus_scatter <- function(x = "Mono.intermediate__Mitochondrial.dependence"){
    
    dt <- read_tsv(file = "../data/correlation_results_signif.tsv", col_names = T, show_col_types = F) 
    all_measure_mat <- readRDS(file = "../data/all_measure_signif.rds")
    colnames(all_measure_mat) <- str_replace_all(colnames(all_measure_mat), pattern = " ", "." )
    #dt$measure1 <- str_remove_all(dt$measure1, pattern = ".*:.")
    #dt$measure2 <- str_remove_all(dt$measure2, pattern = ".*:.")
    lol <- dt %>% filter(measure1 == x | measure2 == x)
    all_measures <- unique(c(lol$measure1, lol$measure2))
   
    all_measure_dt <- all_measure_mat[, all_measures]
    all_measure_dt$sample <- rownames(all_measure_dt)
    all_measure_dt <- all_measure_dt %>% 
        pivot_longer(-sample, names_to = "measure", values_to = "value") %>% 
        left_join(pairwise_FMT_dt %>% select(sample, group = name))
    
    
    plots_corr <- lapply(unique(all_measure_dt$measure),
                         function(m){
                             tmp <- all_measure_dt %>% filter(measure %in% c(x, m)) %>% distinct()
                             tmp <- tmp %>% pivot_wider(id_cols = c(sample, group), 
                                                        values_from = "value", names_from = "measure")
                             
                             p <- ggplot(data = tmp, 
                                         aes(x = get(x), y = get(m) )) + 
                                 geom_point(pch = 21, size = 3, aes(fill=group)) + 
                                 scale_fill_manual(values = c("after_FMT" = "#1f78b4", 
                                                              "donor" = "#b2df8a",  
                                                              "before_FMT" = "#a6cee3")) +
                                 xlab(str_replace(x, "__", replacement = "\n")) + 
                                 ylab(str_replace(m, "__", replacement = "\n")) + 
                                 theme_minimal() + 
                                 theme(legend.position = "none") 
                             return(p)
                             
                         })
    leg <- cowplot::get_legend(plots_corr[[1]] + theme(legend.position = "bottom"))
    cowplot::plot_grid(plotlist = plots_corr,leg)
    
    
}

plot_SCENITH_measure_focus_scatter()
```




## Correlation metabo x taxa in specific targets 

```{r}
target_taxa <- read_lines(file = "../data/FMT_affected_species.txt")
target_metabo <- c("plasma_untarg_2-piperidinone", "feces_untarg_2-piperidinone",
                   "feces_untarg_N-methylproline", "plasma_untarg_N-methylproline",
                   "plasma_untarg_proline", "feces_untarg_proline", "feces_untarg_prolylhydroxyproline")
corr_res_type <- read_tsv(file = "../data/correlation_results_all.tsv", col_names = T, show_col_types = F)

corr_1 <- corr_res_type %>% filter(measure1 %in% c(target_metabo, target_taxa), measure2 %in% c(target_metabo, target_taxa) )
corr_1 <- corr_1 %>% filter(type1 == "species_relab" | type2 == "species_relab")

corr_1$Padj_taxa_metabo <- p.adjust(corr_1$p, method = "BH")

corr_1 <- corr_1 %>% arrange(Padj_taxa_metabo)

write_tsv(corr_1, file = "../data/correlation_metabo_taxa_proline_and_co.tsv", col_names = T)


corr_1 %>% filter(Padj < 0.1) %>% arrange(measure1)

```



# Focused scatterplot of cherry-picked correlations


```{r}
all_measure_mat <- readRDS(file = "../data/all_measure_signif.rds")
all_measure_type <- read_tsv(file = "../data/all_signif_measure_type.tsv", col_names = T, show_col_types = F)
sample_keep <- pairwise_FMT_dt %>% pull(sample) %>% unique()


plot_scatter_focus <- function(measure1 = "Mono.classical__Glucose.dependence", measure2 = "feces_untarg_proline"){
    
    tmp_dt <- all_measure_mat[, c(measure1, measure2)] %>% na.omit()
    tmp_dt$sample <- rownames(tmp_dt)
    tmp_dt <- tmp_dt %>% left_join(pairwise_FMT_dt %>% distinct(name, sample))
    
    ggplot(data = tmp_dt,
           aes(x = .data[[measure1]], y = .data[[measure2]])) + 
        ggpubr::stat_cor() + 
        geom_point(aes( fill = name), pch = 21, size = 3) + 
        scale_fill_manual(values = colormap_group) +
        theme_bw() 
}


plot_scatter_focus()
plot_scatter_focus(measure2 = "feces_untarg_N-methylproline" )
plot_scatter_focus("feces_untarg_N-methylproline", "feces_untarg_proline" )

# other misc
plot_scatter_focus(measure1 = "Dorea_longicatena.t__SGB4581")
plot_scatter_focus(measure1 = "Mono.intermediate__Glucose.dependence" )
plot_scatter_focus(measure2 = "plasma_untarg_2-piperidinone" )
plot_scatter_focus(measure2 = "plasma_untarg_proline")
plot_scatter_focus("Mono.intermediate__Mitochondrial.dependence",  "targeted_metabo__plasma__GCDCA")



```




# Proline derivative focus


```{r}
feces_sample_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN FECES).XLSX", sheet = 3) %>% filter(!is.na(GROUP_ID))
feces_chemical_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN FECES).XLSX", sheet = 2)
feces_log_data <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN FECES).XLSX", sheet = 7)
feces_log_data <- data.frame(feces_log_data %>% select(-PARENT_SAMPLE_NAME), row.names = feces_log_data$PARENT_SAMPLE_NAME)

# Plasma data
plasma_sample_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN PLASMA).XLSX", sheet = 3) %>% filter(!is.na(GROUP_ID))
plasma_chemical_metadata <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN PLASMA).XLSX", sheet = 2)
plasma_log_data <- readxl::read_xlsx(path = "../RAW_metabo_untargeted/HSAE-01-24MD+ DATA TABLES (HUMAN PLASMA).XLSX", sheet = 7)
plasma_log_data <- data.frame(plasma_log_data %>% select(-PARENT_SAMPLE_NAME), row.names = plasma_log_data$PARENT_SAMPLE_NAME)
```


```{r}
# extend to 

#proline_derivatives_list <- c("proline", "N-methylproline", "5-aminovalerate", "2-piperidinone", "prolylhydroxyproline")

plot_dt <- lapply(c("feces", "plasma"), function(SITE){
    print(SITE)
    if(SITE == "feces"){
        metabo_measure <- feces_log_data                                         
        sample_dt <- feces_sample_metadata
        chem_md <- feces_chemical_metadata
        fmt_grp_dt <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")
    } else {
        metabo_measure <- plasma_log_data
        sample_dt <- plasma_sample_metadata
        chem_md <- plasma_chemical_metadata
        fmt_grp_dt <- pairwise_FMT_dt %>% filter(data == "blood_scenith")
    }
    TARGET_ID <- chem_md$CHEM_ID[grepl(chem_md$CHEMICAL_NAME, pattern ="prolin")]
    TARGET_ID2 <- chem_md$CHEM_ID[chem_md$CHEMICAL_NAME %in% c("5-aminovalerate", "2-piperidinone")]
    TARGET_ID <- c(TARGET_ID, TARGET_ID2)
    
    sample_dt <- sample_dt %>%  left_join(fmt_grp_dt %>% select(sample_id = sample, FMT_group, name)) 
    metabo_measure <- metabo_measure[sample_dt$PARENT_SAMPLE_NAME, ] %>% na.omit()
    TARGET_ID <- TARGET_ID[paste0("X", TARGET_ID) %in% colnames(metabo_measure)]
    tmp <- metabo_measure[, paste0("X", TARGET_ID)]
    colnames(tmp) <-  chem_md %>% filter(CHEM_ID %in% TARGET_ID) %>% pull(CHEMICAL_NAME)
    tmp$PARENT_SAMPLE_NAME <- rownames(tmp)
    plot_dt <- tmp %>% left_join(sample_dt %>% select(PARENT_SAMPLE_NAME, name, sample_id)) %>% na.omit() %>%
        pivot_longer(-c(PARENT_SAMPLE_NAME, name, sample_id), values_to = "chem_concentration", names_to = "chem_name")
    plot_dt$bodysite = SITE
    return(plot_dt)
}) %>% do.call(., what = rbind)


plot_dt$name <- factor(plot_dt$name, levels = c("before_FMT",  "donor", "after_FMT"))

ggplot(data = plot_dt,
       aes(x = name, y = chem_concentration)) + 
    geom_boxplot(outlier.shape = NA) + 
    ggpubr::stat_pwc(tip.length = 0, p.adjust.method = "BH", p.adjust.by = "panel", hide.ns = T) +
    geom_jitter(height = 0, width = 0.3)  +
    theme_bw() + ylab("Normalized concentration") + xlab("") +
    facet_grid(cols = vars(chem_name), rows = vars(bodysite), scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90))

# feces only
ggplot(data = plot_dt %>% filter(bodysite == "feces"),
       aes(x = name, y = chem_concentration, colour = name)) + 
    geom_boxplot(outlier.shape = NA) + 
    ggpubr::stat_pwc(tip.length = 0, p.adjust.method = "BH", p.adjust.by = "panel", hide.ns = T) +
    geom_jitter(height = 0, width = 0.3)  +
    theme_bw() + ylab("Normalized concentration") + xlab("") +
    scale_color_manual(values = colormap_group) + 
    facet_wrap(~chem_name, nrow = 2, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90), 
          strip.background = element_blank(), 
          panel.grid.major.x = element_blank(), 
          legend.position = "none")



# test high dorea
high_dorea_samples <- read_lines(file = "../data/dorea_high_samples.txt")
plot_dt <- plot_dt %>% mutate(has_high_dorea = ifelse(sample_id %in% high_dorea_samples, "yes", "no"))

ggplot(data = plot_dt %>% filter(bodysite == "feces"),
       aes(x = name, y = chem_concentration)) + 
    geom_boxplot(outlier.shape = NA) + 
    ggpubr::stat_pwc(tip.length = 0, p.adjust.method = "BH", p.adjust.by = "panel", hide.ns = T) +
    geom_jitter(aes(fill = has_high_dorea),
                height = 0, width = 0.3 , pch = 21, size = 3 )  +
    theme_bw() + ylab("Normalized concentration") + xlab("") +
    facet_wrap(~chem_name, nrow = 3, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90))


```


