---
title: "Analysis scenith data"
author: "Léonard Dubois, Rodrigo Formiga and Harry Sokol"
date: "2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(phyloseq)
library(microViz)
library(viridis)
library(vegan)
library(ggsignif)
library(rstatix)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)

rm(list = ls())
gc()


colormap_group <- c("#b2df8a", "#a6cee3", "#1f78b4") 
names(colormap_group) <- c("HD", "before_FMT", "after_FMT") 

comparisons_grp_rm <- c("pre_post_FMT" = "donor", "pre_donor" = "after_FMT", "post_donor" = "before_FMT")
pairwise_FMT_dt <- read_tsv(file = "../data/FMT_pairing.tsv", col_names = T, show_col_types = F)  %>% 
    mutate(FMT_group = paste0("FMT_grp_", row_number())) %>% 
    pivot_longer(cols = c("donor", "before_FMT", "after_FMT"), values_to = "sample")

```



# Plan

For each type of data (Microbiome, metabo, scenigth, Humann) look for differences and evolution between:
- Pre FMT and Post FMT
- Pre FMT and Healthy donors
- Healthy donors and post FMT
And intersect between these test

Measure the shift (pre post FMT) compared to the mean of Healthy donors


Then look for engraftment of species (% of abundance of species coming from donor)


Build matrix of significant measurements (bacteria, metabolites, SCENITH ) to look for correlation


```{r}

all_measure_mat <- NULL
all_measure_type <- tibble()


```



# Microbiota data


## DATA

```{r}
mpa4_table <- read_tsv(file = "../data/profile_mpa4_vJun23_FMT_Cdiff.tsv", col_names = T, show_col_types = F)

rname <- mpa4_table$clade
mpa4_table <- as.matrix(mpa4_table %>% select(-clade))
rownames(mpa4_table) <- rname

mpa4_table_spc <- mpa4_table[ grepl("s__[A-Za-z_]+$", rownames(mpa4_table) ), ]
rownames(mpa4_table_spc) <- str_remove(string = rownames(mpa4_table_spc), ".*s__")
```


## tests

```{r}

spc_keep <- apply(mpa4_table_spc, 1, function(x) sum(x > 0.05)/ncol(mpa4_table_spc) >= 0.3)
mpa_dt_light <- mpa4_table_spc[spc_keep , ]


median_relab <- lapply(rownames(mpa_dt_light), function(x){
    tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics") %>% distinct(name, sample) %>% filter( sample %in% colnames(mpa_dt_light)) 
    tmp$spc_abundance <- mpa_dt_light[x, tmp$sample]
    res <- tmp %>% group_by(name) %>% 
        summarise(med_relab = quantile(spc_abundance, 0.9)) %>%
        mutate(spc = x)
    return(res)
}) %>% do.call(., what= rbind)



res_microbiota <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             #grp_torm = "donor"
                             tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")  %>% 
                                 filter(name != comparisons_grp_rm[comparison], sample %in% colnames(mpa4_table_spc)) 
                             tmp <- tmp %>% group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                             mpa_dt <- mpa_dt_light[, tmp$sample]
                           
                             res_test <- lapply(rownames(mpa_dt), function(x){
                                 tmp$spc_abundance <- mpa_dt[x, ]
                                 if(var(tmp$spc_abundance) == 0){ return(NULL)}
                                 test <- wilcox_test(data = tmp, formula = spc_abundance ~ name, paired = TRUE)
                                 #effsize <- wilcox_effsize(data = tmp, formula = spc_abundance ~ name, paired = TRUE)
                                 diff_median_relab <- tmp %>% group_by(name) %>% 
                                     summarise(median_relab = median(spc_abundance)) %>% 
                                     pull(median_relab) %>% diff()
                                 return(tibble(spc = x, 
                                               comparison = comparison, 
                                               wilcox_p = test$p, 
                                               diff_median_relab = diff_median_relab))
                                 
                             }) %>% do.call(., what= rbind)
                             res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                             return(res_test)
                         }) %>% do.call(., what= rbind)

```

main visualization

```{r}


tmp <- res_microbiota %>% filter(comparison == "pre_post_FMT") %>% 
    mutate(restored = ifelse(diff_median_relab < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.05, restored, "NS")) %>% 
    select(spc, restored)


tmp2 <- res_microbiota %>% filter(comparison == "post_donor") %>% 
    mutate(type_restoration = ifelse(diff_median_relab < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.05, type_restoration, "normal")) %>%
    select(spc, type_restoration)

    
tmp <- full_join(tmp, tmp2) %>% mutate(FMT_effect = paste(restored, type_restoration)) %>% filter(!grepl(FMT_effect, pattern = "NS"))

plot_dt <- median_relab %>% left_join(tmp) %>% na.omit()


ggplot(data = plot_dt, 
       aes(x = med_relab, y = reorder(spc, med_relab))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("90th percentile of relative abundance (log10)") +
    facet_grid(rows = vars(FMT_effect), 
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank(), 
          axis.text.y = element_text(face = "italic"))


```


```{r}
all_measure_type <- all_measure_type %>% rbind(tibble(measure = unique(plot_dt$spc), type = "species_relab"))
all_measure_mat <- cbind(all_measure_mat, t(mpa4_table_spc[unique(plot_dt$spc), ]))

```


## misc pieces of code



Heatmap visualisation

```{r echo=FALSE, eval=FALSE }

res_microbiota_simple <- res_microbiota_simple %>% 
    group_by(spc) %>% filter(min(Padj) < 0.05) %>%
    mutate(spc = str_remove(pattern = ".*s__", spc)) %>% 
    mutate(diff_median_relab = ifelse(diff_median_relab > 0, 1, -1)) %>%
    mutate(diff_median_relab = ifelse(Padj < 0.05, diff_median_relab, 0))

shift_dt <- res_microbiota_simple %>% 
    select(spc, comparison, diff_median_relab) %>% 
    pivot_wider(names_from = comparison, values_from = diff_median_relab) %>%
    as.data.frame()
rownames(shift_dt) <- shift_dt$spc
shift_dt <- shift_dt %>% select(-spc)


pval_dt <- res_microbiota_simple %>% 
    select(spc, comparison, Padj) %>% 
    pivot_wider(names_from = comparison, values_from = Padj) %>%
    as.data.frame()
rownames(pval_dt) <- pval_dt$spc
pval_dt <- pval_dt %>% select(-spc)

col_fun_HM <- colorRamp2(c("blue", "white", "red"),
                             breaks = c(max(shift_dt), 0, min(shift_dt)))
h <- Heatmap(matrix = shift_dt, 
            col = col_fun_HM,
            na_col = "white",
                 
            width = unit(0.30, "cm") * ncol(shift_dt), 
            height = unit(0.30, "cm") * nrow(shift_dt),
            
            column_title_gp = gpar(fontsize = 12),
            column_title_side = "top", 
            cluster_rows = T, 
            show_row_dend = F,
            cluster_columns = F, 
            show_column_names = T,
            show_row_names = T, row_title = NULL,
            
            border = T,                
            column_gap = unit(1, "mm"), 
            row_names_gp = gpar(fontsize = 9, fontface= "italic"),     
            column_names_gp = gpar(fontsize = 9),
            
            rect_gp = gpar(col = "white", lwd = 0.5),
            show_heatmap_legend =T,
            row_names_side = "left", 
            
            cell_fun = function(j, i, x, y, width, height, fill) {
                if(!is.na(pval_dt[i, j] )){
                    if(pval_dt[i, j] < 0.1){
                        color_star <- ifelse(abs(shift_dt[i, j]) > 0.5, "white", "black")
                        grid.text("*", x, y -height/3, gp = gpar(fontsize = 12, col = color_star))
                    }
                }
            })

draw(h, heatmap_legend_side = "right", )


```







# HUMAnN 3 - Pathways

```{r}
humann3_dt <- read_tsv(file = "../data/HUMAnN3_pathabundance.tsv", col_names = T, show_col_types = F)

humann3_pathway_table <- data.frame(humann3_dt %>% select(-pathway), row.names = humann3_dt %>% pull(pathway), check.names = F )
#colnames(humann3_pathway_table) <- str_replace(colnames(humann3_pathway_table), "\\.", "-")
#colnames(humann3_pathway_table) <- str_extract(colnames(humann3_pathway_table) , pattern = "(HD|Cosmic)_[0-9]+")
pathways_relab <- apply(humann3_pathway_table, 2, function(x) 100*x/sum(x))
```


## Test

```{r}

pathway_median_relab <- lapply(rownames(pathways_relab), function(x){
    tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics") %>% distinct(name, sample) %>% filter( sample %in% colnames(pathways_relab)) 
    tmp$pathway_abundance <- pathways_relab[x, tmp$sample]
    res <- tmp %>% group_by(name) %>% 
        summarise(med_relab = median(pathway_abundance)) %>%
        mutate(pathway = x)
    return(res)
}) %>% do.call(., what= rbind)


pathway_keep <- apply(pathways_relab, 1, function(x) sum(x > 0.05)/ncol(pathways_relab) >= 0.5)
#pathway_keep <- apply(pathways_relab, 1, var) %>% sort() %>% head(n = nrow(pathways_relab) * 0.7) %>% names()
pathways_relab_light <- pathways_relab[pathway_keep , ]



res_pathway <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             #grp_torm = "donor"
                             tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")  %>% 
                                 filter(name != comparisons_grp_rm[comparison], sample %in% colnames(pathways_relab)) 
                             tmp <- tmp %>% group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                             pathway_dt <- pathways_relab_light[, tmp$sample]
                           
                             res_test <- lapply(rownames(pathway_dt), function(x){
                                 tmp$pathway_abundance <- pathway_dt[x, ]
                                 if(var(tmp$pathway_abundance) == 0){ return(NULL)}
                                 test <- wilcox_test(data = tmp, formula = pathway_abundance ~ name, paired = TRUE)
                                 diff_median_relab <- tmp %>% group_by(name) %>% 
                                     summarise(median_relab = median(pathway_abundance)) %>% 
                                     pull(median_relab) %>% diff()
                                 return(tibble(pathway = x, 
                                               comparison = comparison, 
                                               wilcox_p = test$p, 
                                               diff_median_relab = diff_median_relab))
                                 
                             }) %>% do.call(., what= rbind)
                             res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                             return(res_test)
                         }) %>% do.call(., what= rbind)

```




```{r}


tmp <- res_pathway %>% filter(comparison == "pre_post_FMT") %>% 
    mutate(restored = ifelse(diff_median_relab < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.01, restored, "NS")) %>% 
    select(pathway, restored)

tmp2 <- res_pathway %>% filter(comparison == "post_donor") %>% 
    mutate(type_restoration = ifelse(diff_median_relab < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.01, type_restoration, "normal")) %>%
    select(pathway, type_restoration)

tmp <- full_join(tmp, tmp2) %>% mutate(FMT_effect = paste(restored, type_restoration)) %>% filter(!grepl(FMT_effect, pattern = "NS"))

plot_dt <- pathway_median_relab %>% left_join(tmp) %>% na.omit()


ggplot(data = plot_dt, 
       aes(x = med_relab, y = reorder(pathway, med_relab))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("Pathway median relative abundance (log10)") +
    facet_grid(rows = vars(FMT_effect), 
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank(), 
          axis.text.y = element_text(size = 7))


```


```{r}
all_measure_type <- all_measure_type %>% rbind(tibble(measure = unique(plot_dt$pathway), type = "pathway_relab"))
all_measure_mat <- cbind(all_measure_mat, t(pathways_relab_light[unique(plot_dt$pathway), ]))

```



# Metabolites

Careful Cosmic_9 missing from metabo data feces

## DATA

```{r}

metabolomic_dt <- read_tsv(file = "../data/all_metabo_dt.tsv", col_names = T, show_col_types = F) 
#    left_join(pairwise_FMT_dt %>% distinct(FMT_group, sample, name)) %>% as.data.frame()

metabolomic_dt_limits <- read_tsv(file = "../data/all_metabo_dt_limit_detection.tsv", col_names = T, show_col_types = F) 

metabo_type <- read_tsv(file = "../data/metabolites_types.tsv", col_names = T, show_col_types = F)
```



```{r}

res_metabo <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             #print(comparison)
                             lapply(c("feces", "plasma"), function(SITE){
                                 #print(SITE)
                                 metabo_dt <- metabolomic_dt_limits %>%
                                     filter(bodysite == SITE)  %>% select(-bodysite)
                                 metabo_dt <- t(data.frame(metabo_dt, row.names = "sample"))
                                 
                                 if(SITE == "feces"){
                                     tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")                                          
                                 } else {
                                      tmp <- pairwise_FMT_dt %>% filter(data == "blood_scenith")  
                                 }
                                 
                                 tmp <- tmp %>% filter(name != comparisons_grp_rm[comparison]) %>% 
                                     filter(sample %in% colnames(metabo_dt)) %>%
                                     group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                                 metabo_dt <- metabo_dt[, tmp$sample] %>% na.omit()
                                 
                                 res_test <- lapply(rownames(metabo_dt), function(x){
                                     #print(x)
                                     tmp$concentration <- metabo_dt[x, ]
                                     if(var(tmp$concentration) == 0){ return(NULL)}
                                     test <- wilcox_test(data = tmp, formula = concentration ~ name, paired = TRUE)
                                     diff_median_conc <- tmp %>% group_by(name) %>% 
                                         summarise(median_conc = median(concentration)) %>% 
                                         pull(median_conc) %>% diff()
                                     return(tibble(metabolites = x, 
                                                   comparison = comparison, 
                                                   wilcox_p = test$p, 
                                                   diff_median_conc = diff_median_conc, 
                                                   bodysite = SITE))
                                     
                                 }) %>% do.call(., what= rbind)
                                 res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                                 return(res_test)
                             }) %>% do.call(., what = rbind)
                         }) %>% do.call(., what = rbind)



median_conc <-  left_join(pairwise_FMT_dt %>% select(-data, -FMT_group), metabolomic_dt_limits) %>% distinct()
median_conc <- median_conc %>% pivot_longer(-c(name, sample, bodysite), names_to = "metabolites", values_to = "cocentration")
median_conc <- median_conc %>% group_by(name, bodysite, metabolites) %>% summarise(median_conc = median(cocentration)) 
    



```


main visualization

```{r}



tmp <- res_metabo %>% filter(comparison == "pre_post_FMT") %>% 
    group_by(bodysite) %>%
    mutate(restored = ifelse(diff_median_conc < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.05, restored, "NS")) %>% 
    select(metabolites, restored)


tmp2 <- res_metabo %>% filter(comparison == "post_donor") %>% 
    group_by(bodysite) %>%
    mutate(type_restoration = ifelse(diff_median_conc < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.05, type_restoration, "normal")) %>%
    select(metabolites, type_restoration)

    
tmp <- full_join(tmp, tmp2) %>% mutate(FMT_effect = paste(restored, type_restoration)) %>% filter(!grepl(FMT_effect, pattern = "NS"))

plot_dt <- median_conc %>% mutate(metabolites = str_replace_all(string = metabolites, pattern = "-| ", ".")) %>% left_join(tmp) %>% na.omit()


ggplot(data = plot_dt, 
       aes(x = median_conc, y = reorder(metabolites, median_conc))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("Median concentration (log10)") +
    facet_grid(rows = vars(FMT_effect), cols = vars(bodysite),
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank())


```


TODO take plasma resuts as well, but plasma HD are not feces HD

```{r}

metabo_site_list <- plot_dt %>% 
    #filter(bodysite == "feces") %>%
    mutate(metabo_site = paste(bodysite, metabolites, sep = "__")) %>% pull(metabo_site) %>% unique()
# metabo feces and metabo plasma
all_measure_type <- all_measure_type %>% rbind(tibble(measure = metabo_site_list, type = "metabolite"))


feces_dt <- metabolomic_dt_limits %>% filter(bodysite == "feces") %>% select(-bodysite) 
feces_dt <- data.frame(feces_dt %>% select(-sample),row.names = feces_dt$sample )
colnames(feces_dt) <- paste0("feces__", colnames(feces_dt))

plasma_dt <- metabolomic_dt_limits %>% filter(bodysite == "plasma") %>% select(-bodysite)
plasma_dt <- data.frame(plasma_dt %>% select(-sample),row.names = plasma_dt$sample )
colnames(plasma_dt) <- paste0("plasma__", colnames(plasma_dt))

bodysite_dt <- merge(plasma_dt, feces_dt, by = "row.names", all = TRUE)
rownames(bodysite_dt) <- bodysite_dt$Row.names

all_measure_mat <- merge(all_measure_mat, bodysite_dt[, metabo_site_list], by = "row.names", all = TRUE)
rownames(all_measure_mat) <- all_measure_mat$Row.names
all_measure_mat <- all_measure_mat %>% select(-Row.names)

```




# SCENITH

Pairwise test before after

```{r}

scenith_dt <- read_tsv(file = "../data/scenith_data.tsv", col_names = T, show_col_types = F)

test_dt <- scenith_dt %>% 
    pivot_longer(-c(type_measure, group, sample), names_to = "cell_type", values_to = "SCENITH_measure") %>% 
    filter(group != "HD") %>%
    left_join(pairwise_FMT_dt %>% filter(data == "blood_scenith", name != "donor") %>% select(FMT_group, sample)) %>%
    mutate(cell_measure = paste(cell_type, type_measure, sep="__"))

res_test <- lapply(unique(test_dt$cell_measure), function(x){

                    tmp <- test_dt %>% filter(cell_measure == x) %>%
                        group_by(FMT_group) %>% filter(n() ==  2) %>% arrange(FMT_group) %>% ungroup()
                    
                    test <- wilcox_test(data = tmp, formula = SCENITH_measure ~ group, paired = TRUE)
                    effsize <- wilcox_effsize(data = tmp, formula = SCENITH_measure ~ group, paired = TRUE)
                    
                    return(tibble(cell_measure = x, 
                                  wilcox_p = test$p, 
                                  wilcox_effsize = effsize$effsize))
}) %>% do.call(., what = rbind) %>% mutate(wilcox_Padj = p.adjust(wilcox_p, method = "BH"))


res_test %>% filter(wilcox_Padj < 0.05)


```


```{r}

cell_measure_signif <- res_test %>% filter(wilcox_Padj < 0.05) %>% pull(cell_measure)

all_measure_type <- rbind(all_measure_type, tibble(measure = cell_measure_signif, type = "SCENITH_measure"))


scenith_dt_signif <- test_dt %>% filter(cell_measure %in% cell_measure_signif) %>% 
    select(cell_measure, SCENITH_measure, sample) %>% 
    pivot_wider(names_from = cell_measure, values_from = SCENITH_measure) 
scenith_dt_signif <- data.frame(scenith_dt_signif %>% select(-sample), row.names = scenith_dt_signif$sample)

all_measure_mat <- merge(all_measure_mat, scenith_dt_signif, by = "row.names", all = TRUE)
rownames(all_measure_mat) <- all_measure_mat$Row.names
all_measure_mat <- all_measure_mat %>% select(-Row.names)


```


Overall doorbell plot with shift to mean of HD

```{r}
scenith_dt <- read_tsv(file = "../data/scenith_data.tsv", col_names = T, show_col_types = F)

plot_dt <- scenith_dt %>% select(-sample) %>% 
    pivot_longer(-c(type_measure, group), names_to = "cell_type", values_to = "SCENITH_measure") %>%
    group_by(type_measure, group, cell_type) %>% summarise(avg_SCENITH = mean(SCENITH_measure)) %>%
    pivot_wider(id_cols = c(type_measure, cell_type), 
                                   values_from = avg_SCENITH, 
                                   names_from = group, names_prefix = "avg_SCENITH_") %>%
    mutate(shift_after_FMT = avg_SCENITH_HD - avg_SCENITH_after_FMT, 
           shift_before_FMT = avg_SCENITH_HD - avg_SCENITH_before_FMT)


plot_dt %>% select(-avg_SCENITH_HD, -avg_SCENITH_after_FMT, -avg_SCENITH_before_FMT) %>%
    pivot_longer(cols = c(shift_after_FMT, shift_before_FMT), 
                 names_to = "before_after", values_to = "shift") %>%
ggplot(data = ., 
       aes(x = shift, y = reorder(cell_type, shift))) + 
     geom_segment( aes(x=shift, xend=0, y=cell_type, yend=cell_type), color="grey") +
        geom_point(aes(shape = before_after, fill = before_after), size = 3) + 
    geom_vline(xintercept = 0) + 
      scale_shape_manual(values = c("shift_after_FMT" = 21, "shift_before_FMT" = 23)) +
    scale_fill_manual(values = c("shift_after_FMT" = "#1f78b4",  "shift_before_FMT" = "#a6cee3")) +
    theme_bw() + 
    ylab("") + xlab("Average SCENITH measurement") +
    facet_grid(cols = vars(type_measure), scale = "free")

```





```{r}


saveRDS(object = all_measure_mat, file = "../data/all_measure_signif.rds")
write_tsv(all_measure_type, file = "../data/all_signif_measure_type.tsv", col_names = T)

```


# Targeted Correlations

do correlation beetween stuff and visualize as graph
build matrix with pairwise correlation of elements (metabo, spc, scenith)

1 element : 
- scenith, 1 type measure, 
- species abundance
- metabo in bodysite


```{r}
all_measure_mat <- readRDS(file = "../data/all_measure_signif.rds")
all_measure_type <- read_tsv(file = "../data/all_signif_measure_type.tsv", col_names = T, show_col_types = F)
all_measure_type$measure <- str_replace_all(all_measure_type$measure, pattern = "/| ", ".")

```



```{r}
sample_keep <- pairwise_FMT_dt %>% filter(name != "before_FMT") %>% pull(sample)

data_for_corr_test <- all_measure_mat[sample_keep, ]

corr_res <- lapply(combn(colnames(data_for_corr_test), m = 2, simplify = F), 
       function(x){
           tmp_dt <- data_for_corr_test[, x] %>% na.omit()
           res <- cor.test(tmp_dt[, 1], tmp_dt[, 2], method = "spearman")
           return(tibble(measure1 = x[1], 
                         measure2 = x[2], 
                         rho = res$estimate, 
                         p = res$p.value))
       }) %>% do.call(., what = rbind)

corr_res$measure1 <- str_replace_all(corr_res$measure1, pattern = "/| ", ".")
corr_res$measure2 <- str_replace_all(corr_res$measure2, pattern = "/| ", ".")



corr_res_type <- corr_res %>% 
    left_join(all_measure_type %>% select(measure1 = measure, type1 = type)) %>% 
    left_join(all_measure_type %>% select(measure2 = measure, type2 = type)) %>%
    filter(type1 != type2) %>%
    mutate(Padj = p.adjust(p, method = "BH"))



lol = corr_res_type %>% filter(Padj < 0.01)

length(unique(c(lol$measure1, lol$measure2)))

write_tsv(lol, file = "../data/correlation_results_signif.tsv", col_names = T)

```





```{r}
library(igraph)

lol <- read_tsv(file = "../data/correlation_results_signif.tsv", col_names = T, show_col_types = F) 
lol$measure1 <- str_remove_all(lol$measure1, pattern = ".*:.")
lol$measure2 <- str_remove_all(lol$measure2, pattern = ".*:.")

lol <- lol %>% filter(type1 == "SCENITH_measure" | type2 == "SCENITH_measure")

# focus on Monocytes
#lol <- lol %>% filter(grepl(x= measure2, "Mono"))

links <- data.frame(source=lol$measure1,
                    target=lol$measure2,
                    importance=lol$rho )

nodes <- data.frame(rbind(lol %>% select(name = measure1, type = type1), 
                          lol %>% select(name = measure2, type = type2)) %>% 
                        distinct())

g <- graph_from_data_frame(d = links, vertices = nodes , directed = F)
#g <- graph_from_data_frame(d = lol, directed = F)

library(viridis)
coul  <- viridis(length(unique(nodes$type))) 
my_color <- coul[as.numeric(as.factor(V(g)$type))]

col_edge <- ifelse(E(g)$importance > 0, "#2c7bb6", "#d7191c")
 
# Make the plot
plot(g, 
     vertex.color=my_color,
     edge.color  = col_edge,
     vertex.size = 3,
     vertex.label.color="black", 
     vertex.label.cex=0.5, 
     layout=layout_with_fr)



pdf(file = "../figures/SCENITH_measure_graph_corr.pdf", height = 15, width = 20)
plot(g, 
     vertex.color=my_color,
     edge.color  = col_edge,
     vertex.size = 3,
     vertex.label.color="black", 
     vertex.label.cex=0.5, 
     layout=layout_with_fr)
dev.off()





```




