---
title: "Analysis scenith data"
author: "Léonard Dubois, Rodrigo Formiga and Harry Sokol"
date: "2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(phyloseq)
library(microViz)
library(viridis)
library(vegan)
library(ggsignif)
library(rstatix)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)

rm(list = ls())
gc()

colormap_group <- c("#b2df8a", "#a6cee3", "#1f78b4") 
names(colormap_group) <- c("HD", "before_FMT", "after_FMT") 

comparisons_grp_rm <- c("pre_post_FMT" = "donor", "pre_donor" = "after_FMT", "post_donor" = "before_FMT")

```


# Import data


```{r}
scenith_dt <- read_tsv(file = "../data/scenith_data.tsv", col_names = T, show_col_types = F)

pairwise_FMT_dt <- read_tsv(file = "../data/FMT_pairing.tsv", col_names = T, show_col_types = F)  %>% 
    mutate(FMT_group = paste0("FMT_grp_", row_number())) %>% 
    pivot_longer(cols = c("donor", "before_FMT", "after_FMT"), values_to = "sample")

```


# Plan

For each type of data (Microbiome, metabo, scenigth, Humann) look for differences and evolution between:
- Pre FMT and Post FMT
- Pre FMT and Healthy donors
- Healthy donors and post FMT
And intersect between these test

Measure the shift (pre post FMT) compared to the mean of Healthy donors


Then look for engraftment of species (% of abundance of species coming from donor)



# Microbiota data


## DATA

```{r}
mpa4_table <- read_tsv(file = "../data/profile_mpa4_vJun23_FMT_Cdiff.tsv", col_names = T, show_col_types = F)

rname <- mpa4_table$clade
mpa4_table <- as.matrix(mpa4_table %>% select(-clade))
rownames(mpa4_table) <- rname

mpa4_table_spc <- mpa4_table[ grepl("s__[A-Za-z_]+$", rownames(mpa4_table) ), ]
```


## tests

```{r}

spc_keep <- apply(mpa4_table_spc, 1, function(x) sum(x > 0)/ncol(mpa4_table_spc) >= 0.3)
mpa_dt_light <- mpa4_table_spc[spc_keep , ]


median_relab <- lapply(rownames(mpa_dt_light), function(x){
    tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics") %>% distinct(name, sample) %>% filter( sample %in% colnames(mpa_dt_light)) 
    tmp$spc_abundance <- mpa_dt_light[x, tmp$sample]
    res <- tmp %>% group_by(name) %>% 
        summarise(med_relab = quantile(spc_abundance, 0.9)) %>%
        mutate(spc = str_remove(string = x, ".*s__"))
    return(res)
}) %>% do.call(., what= rbind)



res_microbiota <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             #grp_torm = "donor"
                             tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")  %>% 
                                 filter(name != comparisons_grp_rm[comparison], sample %in% colnames(mpa4_table_spc)) 
                             tmp <- tmp %>% group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                             mpa_dt <- mpa_dt_light[, tmp$sample]
                           
                             res_test <- lapply(rownames(mpa_dt), function(x){
                                 tmp$spc_abundance <- mpa_dt[x, ]
                                 if(var(tmp$spc_abundance) == 0){ return(NULL)}
                                 test <- wilcox_test(data = tmp, formula = spc_abundance ~ name, paired = TRUE)
                                 #effsize <- wilcox_effsize(data = tmp, formula = spc_abundance ~ name, paired = TRUE)
                                 diff_median_relab <- tmp %>% group_by(name) %>% 
                                     summarise(median_relab = median(spc_abundance)) %>% 
                                     pull(median_relab) %>% diff()
                                 return(tibble(spc = x, 
                                               comparison = comparison, 
                                               wilcox_p = test$p, 
                                               diff_median_relab = diff_median_relab))
                                 
                             }) %>% do.call(., what= rbind)
                             res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                             return(res_test)
                         }) %>% do.call(., what= rbind)

```

main visualization

```{r}


tmp <- res_microbiota %>% filter(comparison == "pre_post_FMT") %>% 
    mutate(restored = ifelse(diff_median_relab < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.05, restored, "NS")) %>% 
    select(spc, restored)


tmp2 <- res_microbiota %>% filter(comparison == "post_donor") %>% 
    mutate(type_restoration = ifelse(diff_median_relab < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.05, type_restoration, "normal")) %>%
    select(spc, type_restoration)

    
tmp <- full_join(tmp, tmp2) %>% mutate(spc = str_remove(string = spc, ".*s__")) %>% 
    mutate(FMT_effect = paste(restored, type_restoration)) %>% filter(!grepl(FMT_effect, pattern = "NS"))

plot_dt <- median_relab %>% left_join(tmp) %>% na.omit()


ggplot(data = plot_dt, 
       aes(x = med_relab, y = reorder(spc, med_relab))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("90th percentile of relative abundance (log10)") +
    facet_grid(rows = vars(FMT_effect), 
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank(), 
          axis.text.y = element_text(face = "italic"))


```


## misc pieces of code

```{r}

res_microbiota_simple <- res_microbiota %>% mutate(diff_median_relab = ifelse(Padj < 0.05, diff_median_relab, 0)) %>% 
    mutate(shift = ifelse(diff_median_relab > 0, "up", "down")) %>%
    mutate(shift = ifelse(diff_median_relab == 0, "NS", shift))  %>%
    arrange(spc, comparison) %>% group_by(spc) %>% summarise(group_shift = paste(shift, collapse = "__")) %>%
    filter(group_shift != "NS__NS__NS") %>% 
    mutate(spc = str_remove(string = spc, ".*s__")) %>% 
    mutate(FMT_effect = case_when(prel(group_shift) ~ )) 
    

plot_dt <- median_relab %>% left_join(res_microbiota_simple %>% distinct(spc, group_shift)) %>% na.omit()


ggplot(data = plot_dt, 
       aes(x = med_relab, y = reorder(spc, med_relab))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    facet_grid(rows = vars(group_shift), 
               scale = "free", space = "free")
```


Heatmap visualisation

```{r}

res_microbiota_simple <- res_microbiota_simple %>% 
    group_by(spc) %>% filter(min(Padj) < 0.05) %>%
    mutate(spc = str_remove(pattern = ".*s__", spc)) %>% 
    mutate(diff_median_relab = ifelse(diff_median_relab > 0, 1, -1)) %>%
    mutate(diff_median_relab = ifelse(Padj < 0.05, diff_median_relab, 0))

shift_dt <- res_microbiota_simple %>% 
    select(spc, comparison, diff_median_relab) %>% 
    pivot_wider(names_from = comparison, values_from = diff_median_relab) %>%
    as.data.frame()
rownames(shift_dt) <- shift_dt$spc
shift_dt <- shift_dt %>% select(-spc)


pval_dt <- res_microbiota_simple %>% 
    select(spc, comparison, Padj) %>% 
    pivot_wider(names_from = comparison, values_from = Padj) %>%
    as.data.frame()
rownames(pval_dt) <- pval_dt$spc
pval_dt <- pval_dt %>% select(-spc)

col_fun_HM <- colorRamp2(c("blue", "white", "red"),
                             breaks = c(max(shift_dt), 0, min(shift_dt)))
h <- Heatmap(matrix = shift_dt, 
            col = col_fun_HM,
            na_col = "white",
                 
            width = unit(0.30, "cm") * ncol(shift_dt), 
            height = unit(0.30, "cm") * nrow(shift_dt),
            
            column_title_gp = gpar(fontsize = 12),
            column_title_side = "top", 
            cluster_rows = T, 
            show_row_dend = F,
            cluster_columns = F, 
            show_column_names = T,
            show_row_names = T, row_title = NULL,
            
            border = T,                
            column_gap = unit(1, "mm"), 
            row_names_gp = gpar(fontsize = 9, fontface= "italic"),     
            column_names_gp = gpar(fontsize = 9),
            
            rect_gp = gpar(col = "white", lwd = 0.5),
            show_heatmap_legend =T,
            row_names_side = "left", 
            
            cell_fun = function(j, i, x, y, width, height, fill) {
                if(!is.na(pval_dt[i, j] )){
                    if(pval_dt[i, j] < 0.1){
                        color_star <- ifelse(abs(shift_dt[i, j]) > 0.5, "white", "black")
                        grid.text("*", x, y -height/3, gp = gpar(fontsize = 12, col = color_star))
                    }
                }
            })

draw(h, heatmap_legend_side = "right", )


```






# Metabolites

Careful Cosmic_9 missing from metabo data feces

## DATA

```{r}

metabolomic_dt <- read_tsv(file = "../data/all_metabo_dt.tsv", col_names = T, show_col_types = F) 
#    left_join(pairwise_FMT_dt %>% distinct(FMT_group, sample, name)) %>% as.data.frame()

metabolomic_dt_limits <- read_tsv(file = "../data/all_metabo_dt_limit_detection.tsv", col_names = T, show_col_types = F) 

metabo_type <- read_tsv(file = "../data/metabolites_types.tsv", col_names = T, show_col_types = F)
```



```{r}

res_metabo <- lapply(names(comparisons_grp_rm), 
                         function(comparison){
                             #print(comparison)
                             lapply(c("feces", "plasma"), function(SITE){
                                 #print(SITE)
                                 metabo_dt <- metabolomic_dt_limits %>%
                                     filter(bodysite == SITE)  %>% select(-bodysite)
                                 metabo_dt <- t(data.frame(metabo_dt, row.names = "sample"))
                                 
                                 if(SITE == "feces"){
                                     tmp <- pairwise_FMT_dt %>% filter(data == "shotgun_metagenomics")                                          
                                 } else {
                                      tmp <- pairwise_FMT_dt %>% filter(data == "blood_scenith")  
                                 }
                                 
                                 tmp <- tmp %>% filter(name != comparisons_grp_rm[comparison]) %>% 
                                     filter(sample %in% colnames(metabo_dt)) %>%
                                     group_by(FMT_group) %>% filter(n() == 2) %>% ungroup()
                                 metabo_dt <- metabo_dt[, tmp$sample] %>% na.omit()
                                 
                                 res_test <- lapply(rownames(metabo_dt), function(x){
                                     #print(x)
                                     tmp$concentration <- metabo_dt[x, ]
                                     if(var(tmp$concentration) == 0){ return(NULL)}
                                     test <- wilcox_test(data = tmp, formula = concentration ~ name, paired = TRUE)
                                     diff_median_conc <- tmp %>% group_by(name) %>% 
                                         summarise(median_conc = median(concentration)) %>% 
                                         pull(median_conc) %>% diff()
                                     return(tibble(metabolites = x, 
                                                   comparison = comparison, 
                                                   wilcox_p = test$p, 
                                                   diff_median_conc = diff_median_conc, 
                                                   bodysite = SITE))
                                     
                                 }) %>% do.call(., what= rbind)
                                 res_test$Padj <- p.adjust(res_test$wilcox_p, method = "BH")
                                 return(res_test)
                             }) %>% do.call(., what = rbind)
                         }) %>% do.call(., what = rbind)



median_conc <-  left_join(pairwise_FMT_dt %>% select(-data, -FMT_group), metabolomic_dt_limits) %>% distinct()
median_conc <- median_conc %>% pivot_longer(-c(name, sample, bodysite), names_to = "metabolites", values_to = "cocentration")
median_conc <- median_conc %>% group_by(name, bodysite, metabolites) %>% summarise(median_conc = median(cocentration)) 
    



```


main visualization

```{r}



tmp <- res_metabo %>% filter(comparison == "pre_post_FMT") %>% 
    group_by(bodysite) %>%
    mutate(restored = ifelse(diff_median_conc < 0, "restored", "depleted")) %>% 
    mutate(restored = ifelse(Padj < 0.05, restored, "NS")) %>% 
    select(metabolites, restored)


tmp2 <- res_metabo %>% filter(comparison == "post_donor") %>% 
    group_by(bodysite) %>%
    mutate(type_restoration = ifelse(diff_median_conc < 0, "over", "partial")) %>% 
    mutate(type_restoration = ifelse(Padj < 0.05, type_restoration, "normal")) %>%
    select(metabolites, type_restoration)

    
tmp <- full_join(tmp, tmp2) %>% mutate(FMT_effect = paste(restored, type_restoration)) %>% filter(!grepl(FMT_effect, pattern = "NS"))

plot_dt <- median_conc %>% mutate(metabolites = str_replace_all(string = metabolites, pattern = "-| ", ".")) %>% left_join(tmp) %>% na.omit()


ggplot(data = plot_dt, 
       aes(x = median_conc, y = reorder(metabolites, median_conc))) +
    geom_point(aes(shape = name, fill = name), size = 2) + 
    scale_shape_manual(values = c("after_FMT" = 21, "donor" = 22,  "before_FMT" = 23)) +
    scale_fill_manual(values = c("after_FMT" = "#1f78b4", "donor" = "#b2df8a",  "before_FMT" = "#a6cee3")) +
    theme_bw() + scale_x_log10() +
    ylab("") + xlab("Median concentration (log10)") +
    facet_grid(rows = vars(FMT_effect), cols = vars(bodysite),
               scale = "free", space = "free") + 
    theme(strip.text.y.right = element_text(angle = 0), 
          strip.background = element_rect(linetype = "blank",
                                           fill = "white"), 
          legend.position = "bottom", 
          axis.ticks.y = element_blank())


```





